SCC3 acts as the cohesin and inhibits inter-sister chromatids repair during rice meiosis

Version 1

Abstract

Abstract

Cohesin is a multi-subunit protein responsible for holding sister chromatids together during mitosis and meiosis. Each subunit is functionally essential, and their deletion is always lethal. SCC3 is a highly conserved constituent of the cohesin complex. However, the exact mitotic and meiotic functions of SCC3 in rice remains to be elucidated. Here, we found null alleles of SCC3 cause embryo lethality. Only scc3 weak mutants could survive and show vegetative and reproductive defects. Specifically, the replication process of sister chromatids is disturbed in scc3 during interphase both in mitosis and meiosis. Moreover, SCC3 has distinct localization patterns between mitotic and meiotic cells. The numbers of DMC1, RAD51 and ZIP4 foci are significantly decreased in scc3, and ZEP1 displays as an abnormal punctate foci during zygotene. Importantly, the scc3 fails to synapse, but in this case chromosome fragmentation is absent. Thus, SCC3 seems to inhibit inter-sister chromatids repair (ISR), and this process is independent of DMC1-mediated ISR.

Introduction

The correct segregation of sister chromatids is essential for mitotic cell stabilization and to generate haploid gametes during meiosis. The cohesin complex is a conserved multi-subunit protein system that guarantees the proper segregation of sister chromatids in mitosis and across living organisms. This complex is formed by four core subunits, including the structural maintenance of chromosomes (SMC) subunits SMC1 and SMC3, and the sister chromatid cohesin (SCC) subunits SCC1 and SCC3 (Bolanos-Villegas et al., 2017; Moronta-Gines et al., 2019). The cohesin complex forms a ring encircling the sister chromatids and ensures genomic stability during DNA replication (Higashi et al., 2020). In addition, cohesin also participates in the repair of DNA double-strand breaks (DSB), homologous pairing, construction of the synaptonemal complex (SC), orientation of kinetochores, and regulation of gene expression (Nasmyth and Haering, 2009).

Cohesin is established during the S phase, and includes arm and centromeric cohesin. Arm cohesin ensures sister chromatids bond together during mitotic prophase. Centromeric cohesin persists until the metaphase-to-anaphase transition, and is vital to resist the force exerted by spindle microtubules, allowing for the accurate segregation of sister chromatids to opposite poles of the cell. The dissociation of cohesin generally occurs in two steps, including through the dissociation of cohesin on chromosome arms during prometaphase, followed by the degradation of cohesin at the centromeres at the onset of anaphase (Ishiguro and Watanabe, 2007; Watanabe, 2005). This allows sister chromatids to segregate in mitosis.

Meiosis contains a single round of DNA replication, but, in contrast to mitosis, is followed by two successive chromosome divisions. Nonetheless, while the loading mechanism of cohesin is very similar to the analogous process during mitosis, the cohesin degradation during meiosis occurs in two steps: from chromosome arms during the first meiotic division, and from the centromeres during the second meiotic division (Mercier et al., 2015). During meiosis I, homologous chromosomes are linked by chiasmata that separate at metaphase I. At this stage, cohesin dissociates from chromosome arms while the two sister chromatids remain closely together because centromeres cohesin are preserved. Subsequently, the sister chromatids separate at anaphase II due to full cohesin disassociation at the chromosome arms and centromeres. During this process, sister chromatids cohesin is protected by shugoshin from cleavage by separase (Wang et al., 2012; Wang et al., 2011), which determines the correct separation of sister chromatids during meiosis.

Notably, the cohesin complex plays additional functional roles in meiosis. Specifically, cohesin on chromosome arms is crucial for normal progression of meiosis I, while centromeric cohesin plays a more prominent role in meiosis II by geometrically orientating the sister chromatids. During meiosis I, homologous chromosomes are segregated in opposite directions to halve the chromosome number, but sister chromatids remain temporarily mono-oriented and move to the same pole. In meiosis II, sister chromatid pairs become bi-oriented and separate towards opposite spindle poles through similar mechanisms as those observed in mitosis (Schvarzstein et al., 2010). In this process, centromeric cohesin plays a crucial role in establishing kinetochore geometry and thus determining sister kinetochore orientation (Gryaznova et al., 2021). In addition, the cohesin complex impacts the formation of chromosome axial elements (AE) and the assembly of the synaptonemal complex (SC) (Agostinho et al., 2016; Fujiwara et al., 2020; Llano et al., 2012). Besides, the cohesin complex regulates gene expression in eukaryotes, acting as a transcriptional co-activator in humans that is recruited by sequence-specific transcription factors (Casa et al., 2020; Kagey et al., 2010; Lara-Pezzi et al., 2004; Remeseiro et al., 2012). Finally, the cohesin complex is now emerging as a key player in DSBs repair pathway choice and efficiency. In yeast, cohesin and DMC1 may interactively modulate homologous recombination partner choice, with program-appropriate effects (Hong et al., 2013; Phipps and Dubrana, 2022; Sanchez-Moran et al., 2007).

To date, there were many researches on cohesin subunits, which had different functions throughout the cell cycle. In yeast, SCC1 ensures sister chromatid cohesin in mitosis and is associated with chromosomes during the S phase, dissociating during the metaphase-to-anaphase transition. The expression of SCC1 gradually decreases at the beginning of meiosis, whereby SCC1 likely plays a minor role during meiosis (Michaelis et al., 1997). Instead, REC8 is a special meiotic-specific cohesin component that replaces SCC1 in yeast (Klein et al., 1999; Watanabe and Nurse, 1999) and which is expressed in leptotene and decreases after pachytene (Shao et al., 2011). REC8 also plays a crucial role establishing sister kinetochore orientation in yeast, mice and plants (Chelysheva et al., 2005; Ogushi et al., 2021; Toth et al., 2000). In addition, rec8 mutants in flowering plants exhibit sticky chromosomes, which severely affects the localization of other key meiotic proteins (Cai et al., 2003; Chelysheva et al., 2005; Shao et al., 2011). Moreover, SCC3-SCC1 subcomplex is necessary for cohesin loading and the establishment of sister chromatid cohesin in mitosis in yeast (Li et al., 2018a; Orgil et al., 2015). SCC3 also being responsible for maintaining centromeric orientation during meiosis in Arabidopsis (Cai et al., 2003; Chelysheva et al., 2005; Kulemzina et al., 2012). Taken together, these results indicated that cohesin is crucial for the stability of chromosome structure in mitosis and meiosis. However, does cohesin have other functions? such as participating in homologous recombination and DSBs repair process? Also, the mechanisms through which SCC3 regulates the assembly of sister chromatid cohesin in mitosis and its function in meiosis were not well investigated. SCC3 and REC8 are both meiotic cohesin, and what is the difference between them?

Here, we further investigated SCC3 functions, demonstrating its involvement in mitosis and meiosis. We observed the replication process of chromosome 12 during interphase confirming that SCC3 stabilizes chromosome replication process both in mitosis and meiosis, which is crucial for sister chromatids cohesin. Furthermore, we investigated the process of cohesin loading and degradation on chromosomes by using the antibody of SCC3 on chromosomes during mitosis and meiosis. We also show the localization of SCC3 is regulated by REC8. Meanwhile, other meiotic proteins (COM1, DMC1, RAD51, PAIR2 and PAIR3) failed to localize in rec8. That is, REC8 is a fundamental meiotic protein which is prerequisite for the localization of other meiotic proteins. However, the meiotic components (DMC1, RAD51, ZIP4, HEI10 and ZEP1) could be present in scc3 but apparently making no functional contribution to recombination and synapsis indicating that SCC3 participates in homologous pairing, synapsis, recombination progress and CO formation. More importantly, we found meiotic DSBs were well repaired in scc3, probably using sister chromatids as a template, which means SCC3, as a member of cohesin complex, could inhibit inter-sister chromatids repair during meiosis. Finally, we explore different roles of SCC3 and REC8 in meiosis. All findings are integrated into a coherent model.

Results

SCC3 causes both vegetative and reproductive growth defects

We performed a BLAST search in the Rice Genome Annotation Project database (http://rice.plantbiology.msu.edu/cgibin/gbrowse/rice) and found a single hypothetical homologue of SCC3 proteins (A. thaliana SCC3, M. musculus STAG3 and H. sapiens SA1). This candidate protein is encoded by the Os05g0188500 gene and shares the highest similarity with AtSCC3. The full-length cDNA of SCC3 is 3755 nucleotides long and comprises 22 exons and 21 introns. SCC3 contains a 3345-nucleotide-long ORF that encodes a 1116aa protein. In higher eukaryotes, SCC3 is an evolutionary conserved protein that contains a highly conserved STAG domain (140-239aa), as predicted by SMART (Figure S1A). Multiple alignments of full-length SCC3 protein sequences and its homologs in other plants revealed the STAG domain is highly conserved across both monocotyledons and dicotyledons (Figure S1B). Homology modeling indicated SCC3 protein structures are strongly conserved in different species, which also possess similar STAG domains (Figure S2A). Additionally, a phylogenetic tree was constructed with full-length SCC3 amino acid sequences that confirmed this protein is conserved in eukaryotes (Figure S2B). We also did multiple sequence alignments of these proteins used in the evolutionary tree analysis (Figure S3).

A previous study in Arabidopsis (Chelysheva et al., 2005) identified a T-DNA insertion at the boundary between intron 5 and exon 6 of SCC3 that caused lethality, whilst a weak allele was able to survive albeit with some developmental defects. To clarify SCC3 function in rice, we used the CRISPR-Cas9 system to generate three transgenic lines (Figure S4). The first allele (scc3-1) contains a frameshift “GA” deletion in exon 2 and the second allele (scc3-2) contains a frameshift “ACCGA” deletion in exon 11. We tested for allelism between these two scc3 mutants by crossing scc3-1+/- (male) and scc3-2+/- (female). Of the 78 F1 plants, 18 were scc3-1+/-, 26 were scc3-2+/-, 34 were wildtype for both loci. Thus, these two mutants are allelic and heterozygous scc3-1/scc3-2 is lethal. We were unable to isolate homozygous mutants in these two allelic progenies (n=128, 87 plants were scc3-1+/- and 41 were wildtype; n=116, 80 plants were scc3-2+/- and 36 were wildtype). These results suggested that scc3-1 and scc3-2 mutants were both embryo-lethal.

In addition, we obtained a weak transgenic line (scc3) containing a frameshift “T” insertion in exon 19 leading to premature termination of translation (Figure S4). We found the scc3 weak mutant exhibits abnormal vegetative growth (Figures 1A and 1B), and thus proceeded with further investigation of its other phenotypic effects, which revealed scc3 decreases plant height, tiller number and panicle length (Figures 1C and S5B-S5D). To investigate the patterns of SCC3 expression, we employed RT-PCR in various tissues and found the gene is ubiquitously expressed, particularly in roots, leaves and small panicles (Figure S5A). Accordingly, the weak mutant of SCC3 causes severe vegetative growth defects in rice plants.

SCC3 is required for sister chromatid cohesin during mitosis

(A) Comparison between seedlings and root tips in wildtype and scc3.

(B) Comparison between wildtype and scc3 mutant plants.

(C) Comparison between wildtype and scc3 mutant panicles.

(D) Pollen grains stained in 1% I2-KI solution in wildtype and scc3. Bars, 50 μm.

(E) The chromosome behaviors of root tip cells in wildtype and scc3, stained with DAPI. Individual chromosomes are marked by red boxes. The distance between the sister chromatids increased significantly at prometaphase, and the sister chromatids separated in advance at metaphase in scc3. Bars, 5 μm.

(F) FISH analysis of mitotic cells with centromere-specific fluorescently labeled probes in the wildtype and scc3. The distance between centromeres increased significantly in scc3. CentO (red) signals indicate centromeres. Chromosomes were stained with DAPI (blue). Bars, 5μm.

(G) The distance between sister chromatids in different mitotic chromosomes. In normal condition, two sister chromatids form tight stick. In scc3, the distance between chromosome arms and telomeres were increased significantly. Curve diagrams show the distance between the arms and the telomeres as measured by IMAGEJ. Bars, 1 μm.

(H) Graphical representation of the frequency of each type of chromosome morphology. The classification was assigned when >50% chromosomes in a spread displayed the indicated morphology.

To elucidate whether SCC3 is involved in meiosis, we performed cytological observation of anthers stained with 1% I2-KI, and found almost all pollen grains were shrunken and inviable (Figures 1D and S5E). Our results also showed scc3 flowers did not set seeds when pollinated with wild-type pollen, suggesting the mutant was sterile. We further observed heterozygous scc3 plants produced progenies with a segregation ratio of 3:1 (normal: dwarf and sterile), indicating the scc3 mutation is recessive and monogenic. These data suggest that scc3 weak mutation dramatically interferes with plant vegetative and reproductive growth, which may be relevant to mitosis and meiosis.

SCC3 is required for sister chromatid cohesin during mitosis

To determine whether SCC3 is involved in sister chromatid cohesin in mitosis, we observed chromosome behavior in root tip cells of scc3 and wildtype. DNA replicates and folds into an ordered structure during interphase in wildtype plants (Figure 1E). After this, the chromosomes condensed and formed short thread-like chromatids during prophase, which developed into further condensed sister chromatids in prometaphase that were aligned on the equatorial plate before separating to different poles of the cell.

In contrast, scc3 exhibited “X”-shaped chromosomes with distant arms and telomeres in prometaphase (Figures 1E and 1G), indicating a loss of cohesin between sister chromatids. Approximately two-thirds of scc3 root tip cells (70.6%, n=88) contained partially or completely separated sister chromatids (Figure 1H), which were loosely arranged on either side of the equatorial plate during the anaphase-to-telophase transition (Figure 1E), demonstrating the sister chromatids separated in advance as they were pulled to the spindle poles of the cell. But we did not observe any aneuploidy in anaphase, because mitotic tubulin was attached to centromeres of sister chromatids during prometaphase, and sister chromatids were still together at this time in scc3, which guarantee sister chromatids separated equally.

Furthermore, we monitored the distance between centromeres using FISH probes with centromere-specific tandem repetitive sequence CentO in both wildtype and scc3 somatic cells during prometaphase. We found the distance between sister centromeres also dramatically increased in scc3 (Figure 1F, n=25). However, we also noted that, despite the distance between sister chromatids increased in scc3 weak mutants, sisters were still observed side-by-side during prometaphase, suggesting that there may be other cohesin proteins to maintain sister chromatids associated. Our results demonstrated SCC3 is required for cohesin between sister telomeres, centromeres and arms during mitosis.

To explore whether SCC3 affects the replication process of sister chromatids, we performed full-length FISH assays to monitor the dynamic phenotype of chromosome 12 during interphase. In the wildtype, sister chromatids were replicating that could be observed during interphase (Figure 2A). Subsequently, two chromosome clumps formed during prophase and developed into high-order linear chromosomes during prometaphase. During this time, two sister chromatids seemed overlapped and condensed into a short stick. However, in scc3 an obvious chromosomes fragmentation was observed during interphase, which means the replicating chromatids are abnormally organized. During prophase, chromosome 12 were still into a lump and unable to condense into a stick phenotype. In prometaphase, these abnormal chromosomes finally form loosely-connected sister chromatids (Figure 2A). These observations indicated that SCC3 is involved in the replication process of sister chromatids, which may be implicated as a regulator of the genome, influencing chromatin composition and dynamics, and crucially genome organization through folding chromosomes.

SCC3 affects the replication process of sister chromatids during mitosis

(A) The dynamic process of chromosome replication in early mitosis as revealed by pooled oligos specific to chromosome 12 (red). In the wildtype, S1 and S2 indicate the sister chromatids that are replicating. However, in scc3 sister chromatids exhibit many chromosome fragmentation during replication at interphase. Mitotic chromosomes in wildtype and scc3 were stained with DAPI (blue). Bars, 5 μm.

(B) The loading pattern of SCC3 (green, from mouse) and CENH3 (red, from rabbit) in wildtype root tip cells. Chromosomes were stained with DAPI (blue). Bars, 5 μm.

To investigate the localization of SCC3 during mitosis, we generated a mouse polyclonal antibody against C-terminal and performed immunostaining of root tip wildtype cells (Figure 2B). We noted no SCC3 signal was present in scc3 root tip cells, confirming the specificity of the SCC3 antibody utilized in mitosis (Figure S6A). We found SCC3 was uniformly distributed in the cells during interphase in most cases. From interphase to prophase, as chromosome replication completed, the centromere signals became brighter and SCC3 signals also become denser and brighter. SCC3 eventually converged into intact chromosomes at prometaphase. From metaphase to telophase, SCC3 signal disappeared (Figure 2B). These observations suggested that SCC3 may enhance the interaction with DNA during the transition from interphase to prophase, and dissociated from chromosomes after prometaphase.

SCC3 acts as an axial element during meiosis

REC8 is a special axial protein indicating meiotic progression (Shao et al., 2011). To evaluate SCC3 localization in meiosis, we performed dual-immunolocalization between SCC3 and REC8 in wildtype meiocytes. REC8 signals were initially detected as foci in the nucleus at leptotene, while SCC3 was detectable simultaneously (Figure 3A). From leptotene to zygotene, SCC3 signals were detected as similar elongated foci that formed filamentous structures on the chromosomes. During pachytene and diplotene, constant SCC3 signals were distributed along chromosome axes (Figure 3A). At diakinesis, axial SCC3 signals became diffuse and centromeric signals gradually became clear. At metaphase I, only centromeres SCC3 antibody signals remained (Figure 3B). However, REC8 signals were difficult to detect from diakinesis to metaphase I (Figure S7). In order to further clarify the localization pattern between SCC3 and REC8, we performed super-resolution imaging (SIM) on wildtype meiocytes, which showed REC8 and SCC3 were parallel and colocalized forming parallel axial elements (Figure 3C). In addition, central elements ZEP1 formed two parallel lines surrounded by SCC3 signals (Figure 3D). We also noted no SCC3 signal was present in scc3 meiocytes, confirming the specificity of the SCC3 antibody utilized in meiosis (Figure S6B).

SCC3 is an axial element during meiosis

(A) In meiosis, SCC3 (green, from mouse) colocalizes with REC8 (red, from rabbit) from the leptotene to diplotene. Bars, 5 μm.

(B) During diakinesis and metaphase I, SCC3 (green, from mouse) gradually dispersed and finally retained in the centromeres indicated by CENH3 (red, from rabbit). Chromosomes were stained with DAPI (blue). Bars, 5 μm.

(C) Immunostaining of SCC3 (cyan, from mouse) and REC8 (red, from rabbit) in wildtype meiocytes at pachytene. Two parallel linear SCC3 signals colocalize with the REC8 linear signals, indicating chromosomal axial elements. Magnified images of the blocked regions are shown on the right. Bars, 5 μm.

(D) Immunostaining of SCC3 (cyan, from mouse) and ZEP1 (red, from rabbit) in wildtype meiocytes at pachytene. Two linear SCC3 signals wrap the ZEP1 signals, indicating chromosomal central elements are wrapped by axial elements. Magnified images of the blocked regions are shown on the right. Bars, 5 μm.

SCC3 is required for sister chromatid cohesin in early meiosis

The interphase of meiosis is similar to mitosis, and sister chromatids replicate during interphase and are linked together by cohesin. To determine whether SCC3 is involved in sister chromatid cohesin during meiosis, we observed chromosome behaviors of meiocytes in scc3 and wildtype before leptotene. Cytological phenotypes of meiosis initiation in rice have been reported in detail (Zhao et al., 2018). The first meiotic stage that can be distinguished is preleptotene, and sister chromatids should have completed replication during this time. In preleptotene, the meiocytes chromosomes were discrete and partially contracted during preleptotene (Figure 4A), and subsequently formed thin threads in the wildtype at leptotene. In scc3, chromosomes at preleptotene exhibited dispersed appearance with blurred outlines. During leptotene, chromosomes maintained a loose shape and no filamentous threads were present. These observations indicate SCC3 participates in sister chromatids replication process in early meiosis.

SCC3 is required for sister chromatid cohesin in early meiosis

(A) Meiocyte chromosome morphology at the preleptotene and leptotene in wildtype, scc3, rec8 and rec8 scc3. Chromosomes at preleptotene exhibited a hairy rodlike appearance and formed thin threads at leptotene in wildtype. However, chromosomes in scc3 were dispersed with blurred outlines and failed to from thin threads at leptotene. Bars, 5 μm.

(B) The process of chromosome replication in preleptotene is revealed by pooled oligos specific to chromosome 12 (red). Preleptotene chromosomes in wildtype, scc3, rec8 and rec8 scc3. Chromosomes were stained with DAPI (blue). Bars, 5 μm.

We then examined chromosome behavior in rec8 single mutants and scc3 rec8 double mutants at preleptotene (Figure 4A). Unexpectedly, rec8 exhibited elongated chromosomes similar to the wildtype, whereas scc3 rec8 exhibited diffused chromosomes similar to scc3. These results indicate REC8 does not affect the early replication process of sister chromatids, and the chromosome defects in scc3 rec8 double mutant may be caused by the mutation of SCC3. In addition, we performed full-length FISH assays to monitor chromosome 12 at preleptotene (Figure 4B), and found that replicated chromosome 12 can form short rod-shaped threads in both wildtype and rec8, but appears as a diffuse structure in scc3 and scc3 rec8 mutants due to abnormal replication process. These observations are consistent with DAPI staining results, further demonstrating SCC3 affects sister chromatid replication process in early meiosis.

Homologous chromosome pairing and synapsis are disturbed in scc3

To study the causes of sterility in scc3, we investigated male meiotic chromosome behavior of wildtype and scc3 meiocytes. In the former, homologous chromosomes paired at early zygotene and fully synapsed at pachytene. Moreover, during metaphase I, 12 bivalents were orderly arranged on the equatorial plate, with homologous chromosomes separating at anaphase I (Figure 5A). In contrast, scc3 chromosome behavior showed no differences at zygotene, with meiotic defects becoming apparent at pachytene, when most chromosomes dispersed as single threads with incomplete homologous paring (Figure 5A). We identified 24 univalents during diakinesis and metaphase I (n=161, 150 meiocytes have 24 univalents, 11 meiocytes have 1-2 bivalents), which were always pulled to opposite poles leading to sister chromatids separation at anaphase I. However, in spo11 mutants containing 24 univalents, sister chromatids were separated randomly or still linked together (Figure S8A). These observations show SCC3 determine the separation of sister chromatids during anaphase I, and is involved in homologous pairing, which could influence the normal process of synapsis.

Homologous pairing and synapsis are disturbed in scc3

(A) Meiotic chromosome behavior in wildtype and scc3. Bars, 5 μm.

(B) Immunolocalization of ZEP1 (green, from mouse) and REC8 (red, from rabbit) in wildtype and scc3 meiocytes. ZEP1 was severely suppressed from zygotene to late pachytene in scc3. Bars, 5 μm.

The telomere bouquet is a prerequisite for homologous pairing at early zygotene (Zhang et al., 2020). We performed a FISH assay to detect bouquet formation in wildtype and scc3 meiocytes using a telomere-specific probe (pAtT4). As expected, the wildtype telomeres were attached to a confined region, forming a typical bouquet on the nuclear envelope at zygotene. Interestingly, the formation of telomere bouquets also occurred normally in scc3 (Figure S8B). We then used 5S rDNA to monitor homologous pairing, and detected two overlapping 5S rDNA foci at pachytene in the wildtype, but two separated 5S rDNA foci in unpaired homologous chromosomes in scc3 (n=52; Figure S8C), confirming SCC3’s role in homologous pairing.

Homologous chromosomes align and then become closely juxtaposed by a proteinaceous structure known as the synaptonemal complex (SC), which is comprised of two parallel arrays of axial elements (AE), termed lateral elements (LE) prior to synapsis, that form the structural basis for the assembly of the SC (Ren et al., 2021). Since SCC3 is a LE component, we monitored the assembly of the SC through immunostaining using an antibody against ZEP1 in scc3 and wildtype meiocytes. In the wildtype, we observed ZEP1 gradually elongated and formed short linear signals during zygotene, which eventually extended along the paired homologous chromosomes at pachytene (Figure 5B). However, scc3 mutants displayed ZEP1 as a few punctate foci during zygotene. These abnormal structures were also present at pachytene (Figure 5B), suggesting SCC3 plays a crucial role in the assembly of the SC. In addition, we also examined the localization of other LEs (PAIR2 and PAIR3), and these proteins behave normally in scc3 (Figure S9).

SCC3 is essential for recombination progress and CO formation

Since homologous pairing was impaired in scc3 meiocytes, we wondered whether recombination was normal in scc3. γH2AX is a marker that is used as a proxy to monitor DNA DSBs. Therefore, we investigated the number of γH2AX foci to explore the DSB formation. There was no significant difference in the number of γH2AX foci between the wildtype and scc3 (265.4 ± 8.6, n=31 for scc3 meiocytes; 262.2 ± 7.2, n=20 for wildtype meiocytes) (Figure 6A). Additionally, we also examined the DSB resection factor (COM1), meiotic strand-invasion and exchange factors (RAD51, DMC1) and the Class I, interference-sensitive crossover (CO) marker (ZIP4) in both wildtype and scc3. Immunostaining results revealed that the number of COM1 foci (290.3 ± 8.8, n=26 for scc3 meiocytes; 293.8 ± 9.4, n=23 for wildtype meiocytes) were not significantly different between the wildtype and scc3 (Figures 6A and 6B). However, DMC1 (65.4 ± 17.1, n=25 for scc3 meiocytes; 281.2 ± 8.3, n=20 for wildtype meiocytes), RAD51 (55 ± 21.7, n=26 for scc3 meiocytes; 306 ± 11.9; n=24 for wildtype meiocytes) and ZIP4 (58 ± 17.9, n=23 for scc3 meiocytes; 316.6 ± 11.1, n=21 for wildtype meiocytes) foci were significantly decreased in scc3, suggesting that recombination progress and CO formation was disturbed in scc3. Previous studies revealed that HEI10 is a marker of Class I COs. In wide type, prominent HEI10 foci were observed (24.2 ± 3.8, n=25) at late pachytene (Figures 6A and 6B). By comparison, in scc3 most HEI10 foci were sparse and did not present as large bright signals. Taken together, these results suggested that CO maturation was disturbed in scc3. Although recombinant proteins could be localized in a small portion, the homologous recombination process was still greatly restricted.

SCC3 affects recombination progress and CO formation

(A) Immunolocalization of histone γH2AX phosphorylation (red), COM1 (green, from mouse), DMC1 (green, from mouse), RAD51 (green, from rabbit), ZIP4 (green, from mouse) and HEI10 (green, from mouse) in wildtype and scc3. REC8 (red, from mouse and rabbit) signals were used to visualize chromosome axes. Bars, 5 μm.

(B) Box scatter plot of histone γH2AX phosphorylation, COM1, DMC1, RAD51 and ZIP4 in wildtype and scc3. No difference of histone γH2AX phosphorylation and COM1 were shown between the wildtype and scc3. DMC1, RAD51 and ZIP4 foci were significantly decreased in scc3 compared with wildtype. Values are means ± SD. *** represents P < 0.001, two-tailed Student’s t-tests was performed.

(C) Chromosome behavior in scc3, rec8, rec8 scc3, dmc1 and dmc1 scc3 from diakinesis to anaphase I. Bars, 5 μm.

SCC3 inhibits inter-sister chromatids repair during meiosis

It is worth noting that DSBs formation was normal in scc3, and inter-homolog strand-invasion is destabilized which leads to the defects in CO formation. Nevertheless, in scc3, the meiotic DSBs are reliably repaired as there is no chromosome fragmentation, most likely using sister chromatid as a repair template, leading to 24 univalents (Figures 5A and 6C). These observations indicated that SCC3, as a member of cohesin complex, had the function of inhibiting inter-sister chromatids exchange during meiosis. We also observed chromosome behaviors of rec8 and rec8 scc3 double mutants from diakinesis to anaphase I (Figure 6C). Both of these mutants contained defective chromosomes displaying a highly sticky phenotype from diakinesis to metaphase I, which is similar to the phenotype of non-homolog end joining (NHEJ). Moreover, chromosome fragmentation appeared in anaphase I, illustrating that meiotic DSBs were not well-repaired in rec8 and rec8 scc3. In addition, DMC1 is also responsible for inter-sister exchange during meiosis in both arabidopsis and rice(Kurzbauer et al., 2012; Wang et al., 2016). To test whether DMC1 affects inter-sister repair process in scc3, we generated dmc1 scc3 double mutant. The meiotic defects of dmc1 and dmc1 scc3 were identical from diakinesis to metaphase I, leading to 24 univalents (Figure 6C). Compared with dmc1, the sister chromatids of dmc1 scc3 were completely separated in anaphase I and no chromosome fragmentation were observed. Taken together, meiotic DSBs in scc3 is probably repaired by using sister chromatid as templates, and the inter-sister chromatids exchange in scc3 is independent of DMC1.

The loading of SCC3 onto meiotic chromosomes depends on REC8

After observing that SCC3 is an axial element in meiosis, we evaluated whether its localization can be affected by other meiotic cohesin protein, specially REC8. Firstly, we investigated the number of γH2AX foci to explore the DSB formation in rec8, and found that the DSBs formation was normal in rec8 (Figure 7A). In addition, DSBs ends resection factor COM1 (n=25 for 3 rec8 mutants) and inter-homologue strand-invasion factor RAD51 (n=26 for 3 rec8 mutants) fail to localize in rec8 (Figure 7A). Besides, we detected the distribution of SCC3 in rec8. we found no SCC3 signals in rec8 mutants, suggesting REC8 affects the normal progression of SCC3 loading (n=36 for 3 rec8 mutants). PAIR2 (n=28 for 3 rec8 mutants) could only form short, interrupted stretches and few PAIR3 signals were detectable (n=32 for 3 rec8 mutants). These results suggested that meiotic DSBs formation was normal in rec8, but DSBs ends resection and homologous recombination process were significantly defective. Thus, we speculated that the highly sticky chromosomes and dissociative chromosomes fragmentation from diakinesis to anaphase I in rec8 was caused by COM1-mediated NHEJ pathway.

SCC3 loading onto meiotic chromosomes depends on REC8

(A) Immunolocalization of γH2AX (red, from rabbit), COM1 (green, from mouse), RAD51 (green, from rabbit), SCC3 (green, from mouse), PAIR2 (green, from rabbit) and PAIR3 (green, from rabbit) in rec8 meiocytes at zygotene. CENH3 (red and green, from rabbit and mouse) was used to indicate the centromeres. Bars, 5 μm.

(B) Immunolocalization of REC8 (red, from rabbit) and SCC3 (green, from mouse) in spo11, com1, dmc1, zep1, pair2 and pair3 meiocytes at zygotene. Bars, 5 μm.

(C) SCC3 interacts with RAD21.1 in yeast-two-hybrid assays. Interactions between bait and prey were examined on SD/DDO (SD-Leu-Trp) control media and SD/QDO (SD-Ade-His-Leu-Trp) selective media. AD, prey vector; BD, bait vector.

(D) Bimolecular fluorescence complementation assays show the interactions between SCC3 and RAD21.1 in rice protoplasts. Bars, 5 μm.

PAIR2 and PAIR3 localize on chromosome axes in rice, which is essential for the formation of synaptonemal complex installation. We further investigated the distribution of SCC3 in pair2 and pair3 mutants (Figure 7B). Specifically, SCC3 signals were normal in pair2 and pair3 mutants during meiosis. Moreover, SCC3 distribution was normal in other mutants that affect meiotic DSB formation (spo11-1), strand invasion and homology searching (com1 and dmc1), and synaptonemal complex formation (zep1) (Figure 7B), indicating SCC3 localization is independent of the recombination process.

SCC3 interacts with RAD21.1

The cohesin complex is a ring-shaped protein complex where subunits interact with each other (Li et al., 2018a; Murayama and Uhlmann, 2015; Roig et al., 2014). To confirm this, we performed Y2H assays between several cohesin proteins. Surprisingly, we found SCC3 only interacts with RAD21.1 (Figure 7C), which was further confirmed in BiFC assays (Figure 7D). However, it is worth noting that SCC3 could not interact with other cohesin proteins, including REC8, SMC1 and SMC3 (Figure 7C), indicating that SCC3 could function directly in combination with RAD21.1 in the cohesin complex.

Previous studies showed cohesin regulates genes expression (Cuadrado et al., 2015; Gligoris and Lowe, 2016; Remeseiro et al., 2012), we therefore investigated expression levels of cohesin-related genes in wildtype and scc3 spikelets using RT-qPCR. Interestingly, we found cohesin subunits SCC1 and REC8, as well as centromeric cohesin protector SGO1 were significantly downregulated in scc3 (Figure S10A). In contrast, the cohesin subunits SMC1 and SMC3, as well as lateral/central elements PAIR3 and ZEP1 showed no differences in gene expression. The only upregulated axial element in scc3 spikelets was PAIR2.

We then investigated cohesin gene expression levels in axial/lateral elements mutants (pair2 and pair3) (Figure S10B), and found SMC3, SCC3 and SCC1 were significantly downregulated compared to the wildtype. These findings suggested that SCC3 may thus be involved in the transcriptional regulation of meiosis genes, while axial elements affect cohesin proteins by regulating the expression levels of associated genes.

Discussion

Chromosome segregation during mitosis and meiosis includes a series of highly organized events, such as the formation of a cohesin complex containing different subunits that play multiple functions (e.g., DSB repair, or gene expression regulation) (Moronta-Gines et al., 2019; Nasmyth and Haering, 2009). Here, we show SCC3 is a conserved cohesin subunit affecting vegetative growth and sister chromatid assembly in both mitosis and meiosis. Crucially, our results also indicate SCC3 is a component of LE/AE that is required for homologous pairing and synapsis during meiosis, which is essential for the recombination process and CO formation. More importantly, SCC3, as a member of cohesin complex, may inhibit inter-sister chromatids repair during meiosis, which provides new insights between cohesin and inter-sister/homolog recombination partner choice in meiotic DSBs repair.

Current cohesin complex research mainly focuses on mitosis because the morphology of sister chromatids during meiosis is difficult to observe. In human cells, sister chromatids separate in mitosis in the absence of the SCC3 homologues SA1 and SA2. The former is required for telomere and arm cohesin, whereas the latter is essential for centromere cohesin (Canudas and Smith, 2009). In rice, SCC3 has only a single copy, therefore we observed these all kinds of defects in scc3. SCC3 is not only essential for the stability of sister chromatids cohesin but also plays a role in loading cohesin onto chromosomes. In yeast scc3, the cohesin trimer composed of SCC1, SMC1 and SMC3 remains intact. However, the trimer fails to localize to chromosomes, probably because SCC3 helps maintaining the stability of cohesin (Orgil et al., 2015; Pathania et al., 2021). Additionally, SCC3 contains a conserved DNA binding domain in the SCC3-SCC1 subcomplex, which is required for cohesin association to chromatin (Higashi et al., 2020). In yeast, the SCC3-SCC1 DNA-binding region recruits cohesin complexes to chromosomes and facilitates the function of cohesin during mitotic cell division (Kulemzina et al., 2012; Li et al., 2018a). These studies implied that SCC3-SCC1 interaction is necessary for cohesin ring stability and for maintaining its association with DNA. In Arabidopsis, SCC3 is also involved in vegetative growth of plants, but its role in mitosis remains unclear (Chelysheva et al., 2005). Building from these previous observations, our study demonstrated that SCC3 is encoded by a single gene that participates in early cohesin formation in mitosis. In scc3, vegetative growth was severely inhibited. We found that the replication process of sister chromatids during interphase was abnormal in scc3, possibly because SCC3 is required for cohesin to stably bind to the chromosomes. Also, we observed an increased distance between sister chromatids in prometaphase, but sister chromatids remain partially dissociated not a complete separation, indicating other cohesin subunits are still functional. Taken together, our results suggest SCC3 is a conserved cohesin protein that plays a very important role in mitosis.

SCC3 is also involved in meiotic chromosome structure and plays a crucial role in homologous recombination and inter-sister chromatid repair during meiosis. In vertebrates, SCC3 is encoded by two genes paralogues, including the meiosis-specific subunit STAG3. In human stag3, SIM analysis showed the chromosome axes are regionally separated and the synaptonemal complex is largely reduced (Biswas et al., 2016; Fukuda et al., 2014; Hopkins et al., 2014; Llano et al., 2012). These observations suggest STAG3 is required for synapsis between homologous chromosomes. Compared to vertebrate cells, the morphology of meiocytes is more readily observed during plant meiosis. In Arabidopsis scc3, a total loss of fertility is associated with SCC3’s role in meiosis, with unstable homologous pairing at pachytene. SCC3 is also necessary for monopolar orientation of kinetochores at meiosis I and the maintenance of centromeric cohesin at anaphase I (Chelysheva et al., 2005). In other cohesin mutants, Atscc2 is required for loading of the meiosis-specific cohesin subunit REC8 (Wang et al., 2020), suggesting defects in cohesin-related genes might impact the ability of the cohesin complex to bind chromosomes. Accordingly, in meiosis, we found rice SCC3 has functions in the maintenance of replication process of sister chromatids in preleptotene. SCC3 localizes between homologous chromosomes as an important axial element during meiosis, ensuring correct homologous pairing and synapsis. Meanwhile, SCC3 is required for normal localization of DMC1, RAD51, ZIP4 and HEI10, which is essential for recombination progress and CO formation. Our working model proposes SCC3 is a part of the cohesin ring, and functions in specific association with SCC1 (Figures 8A). Considering SCC3’s role in cohesin and multiple meiotic functions, we speculate that SCC3 may act as LE to stabilize loop boundaries to promote sister chromatids cohesin during meiosis (Figure 8B). More importantly, meiotic homologous pairing and synapsis are defective in scc3, and 24 univalents could be observed at diakinesis and sister chromatids are completely separated at anaphase I with no chromosome fragmentation, which illustrates that meiotic DSBs are well repaired and could only use sister chromatids as a template to repair. Known meiotic components (COM1, DMC1 and RAD51) can be present in scc3 but apparently making no functional contribution to homologous recombination, but these proteins could also have the effect of creating a leading end that can search for an available inter-sister partner, since DMC1 and RAD51 are also involved in inter-sister chromatids repair process. We propose that, when cohesin is present, meiotic recombination exhibits inter-homolog bias to repair. Nonetheless, when cohesin is absent and homologous pairing is defective, recombination exhibits inter-sister bias to repair (Figure 8C). These observations imply that SCC3 acts as a component of cohesin complex, which could block inter-sister chromatids exchange during meiosis.

SCC3 acts as the cohesin and inhibits inter-sister chromatids repair during rice meiosis

(A) Structure model of the cohesin complex. The SCC3 subunit is associated with the middle region of SCC1.

(B) During meiosis, SCC3’s localization pattern probably changes to the root of DNA loop with lateral elements (LE), which wraps synaptonemal complex (SC). This indicates SCC3 is a meiotic LE essential for homologous chromosome pairing and synapsis, also affects recombination progress and CO formation.

(C) Meiotic recombination exhibits an intrinsic and mechanistical default towards selection of homolog bias in wildtype. Cohesin is responsible for the connection of sister chromatids intervening in this mechanism to inhibit recombination into use of sisters. However, in scc3, sister bias is required, with the defects of homologous pairing and synapsis. This outcome may be achieved by the residue function of DMC1 and RAD51 promoting DSBs end association with sister. While this situation in rec8 is quite different with most meiotic proteins fail to localize, thus forced to NHEJ repair pathway causing the sticky chromosomes and fragmentation.

Cohesin is a multi-protein complex that plays a vital role establishing sister chromatid cohesin and high-order chromosome architecture in somatic and germ cells. However, distinct meiosis-specific subunits likely perform different functions (Cuadrado et al., 2015; Ishiguro, 2019; Schvarzstein et al., 2010). Our observations suggest that SCC3 and REC8 may play different roles during mitosis and meiosis. REC8 was considered a meiosis-specific protein, which played the role of SCC1 in meiosis to form cohesin complex. In this study, we found REC8 and SCC3 localization patterns were similar but not identical. Notably, REC8 was not present in mitotic cells. However, we found SCC3 is active in mitosis and meiosis, and may be involved in cohesin loading of the replication process of sister chromatids at preleptotene. It’s worth noting that the mechanism of loading cohesin onto chromosomes in mitosis and meiosis was the same. Meiotic cohesin loaded onto chromosomes during interphase, probably when the sporogenous cell was replicating and acquired meiotic fate. This indicates REC8 is not involved in establishing cohesin at this instance. We also observed SCC3 signals at centromeres from diakinesis to metaphase I, but could not detect any REC8 at metaphase I. These different localization patterns also suggested they may have different functions in meiosis. Previous studies on Arabidopsis rec8 mutants showed sticky chromosomes, and it was difficult to explain the reason of the phenotype (Cai et al., 2003; Lambing et al., 2020). Correspondingly, rice rec8 meiocytes also contained defective chromosomes displaying a highly sticky phenotype from the diakinesis to metaphase I (Shao et al., 2011). Here, we found the chromosomes in scc3 was not sticky as observed in rec8 (Figure 6C). Meanwhile, the majority of meiotic proteins (including SCC3) failed to localize in rec8. This indicates the normal localization of SCC3 and other meiotic proteins depends on REC8, which means REC8 may be a fundamental meiotic protein that is the prerequisite for other meiotic proteins loading. Also, the DSBs end processing and inter-sister/homolog strand-invasion are totally defective in rec8. Thus, we speculated that the sticky chromosomes in rec8 may be caused by the NHEJ repair pathway (Figure 8C). The presented results provide that SCC3 acts as cohesin and stabilizes sister chromatid replication process in both mitosis and meiosis, but REC8 only functions in meiosis. Both proteins are independent of each other in mitosis, but related in meiosis, with REC8 representing a more important player in meiosis.

Material and methods

Meiotic chromosome preparation

Young rice panicles were fixed in Carnoy’s solution (ethanol: acetic acid, 3: 1, v/v). Microsporocytes at different meiotic stages were squashed in an acetocarmine solution on glass slides. Moreover, the slides were covered with a coverslip and washed with 45% acetic acid from each side. Subsequently, the slides were frozen in liquid nitrogen for 10 seconds, and then the coverslip was removed as soon as possible ensuring sufficient cells were fixed on the slide. After ethanol gradient dehydration (70%, 90% and 100%), the slides with chromosome spreads were counterstained with 4’,6-diamidino-2-phenylindole (DAPI, Vector Laboratories, Burlingame, CA, USA) solution. Chromosome behaviors were observed and captured using a ZEISS A2 microscope imaging system with a micro-charge-coupled (micro CCD) device camera.

Antibody production

To generate the antibody against SCC3, a 648 bp C-terminal fragment of SCC3 complementary DNA (amino acids 900-1116) was amplified using primers SCC3-Ab-F and SCC3-Ab-R (Table S1). The PCR product was cloned into the expression vector pET-30a (Novagen, Madison, WI, USA). The fusion peptide expression and purification were carried out as described previously (Miao et al., 2021). The polyclonal antibody was raised from mouse and rabbit. Other antibodies were generated in our laboratory previously.

Immunofluorescence assay

Fresh young panicles were fixed in 4% (w/v) paraformaldehyde for 30 min at room temperature. Different meiotic stages of anthers were squashed in a drop of 1×PBS solution on glass slides. After freezing in liquid nitrogen and ethanol dehydration, slides were incubated with different combinations of diluted antibodies (1:100) in a humid chamber at 37 °C for 2 h. Thereafter, the slides were washed with 1×PBS solution three times and were further incubated for 1 h with the appropriate fluorochrome-coupled secondary antibody, including fluorescein-isothiocyanate-conjugated goat anti-mouse antibody (Southern Biotech, Birmingham, AL, USA), rhodamine-conjugated goat anti-rabbit antibody (Southern Biotech), and AMCA-conjugated goat anti-guinea pig antibody (Jackson ImmunoResearch, West Grove, PA, USA). Slides were washed with 1×PBS solution three times and eventually stained with DAPI. The immunofluorescence signals were analyzed and captured using a ZEISS A2 microscope imaging system.

Fluorescence in situ hybridization

Young panicles of both wild-type and mutants were fixated in Carnoy’s solution (ethanol: acetic acid, 3: 1, v/v). The fluorescence in situ hybridization (FISH) analysis was performed as a detailed protocol previously (Cheng, 2013). The pAtT4 clone containing telomere repeats, the pTa794 clone containing 5S ribosomal RNA genes, and the bulked oligonucleotide probes specific to chromosome 12 were used as probes in FISH analysis. Rhodamine anti-digoxigenin was used for digoxigenin-labeled probes. Chromosome images were captured using an Olympus (Shinjuku-ku, Tokyo, Japan) BX51 fluorescence microscope with a micro CCD camera using software IPLAB4.0 (BD Biosciences, San Jose, CA, USA).

Yeast two-hybrid assay

The yeast two-hybrid (Y2H) assays were conducted by the full-length CDSs of RAD21.1, SCC3, REC8, SMC1 and SMC3 independently cloned into the pGBKT7 or pGADT7 vector. Co-transformants were selected on SD/DDO (SD-Leu-Trp) medium at 30 °C for 3-4 days. Positive interactions of transformants were selected on SD/QDO (SD-Leu-Trp-His-Ade) medium containing aureobasidin A. All primers used to construct plasmids are listed in Table S1. The detailed protocol was described in the manufacturer’s handbook (Yeast protocols handbook; PT3024-1; Clontech).

BiFC assay

To conduct BiFC assays, SCC3 and RAD21.1 were amplified by KOD-plus polymerase and ligated into BiFC vectors, including pSCYNE (SCN) and pSCYCE (SCC). The constructed plasmids were transformed into protoplasts extracted from the young stem of rice etiolated seedlings. After incubation in the dark for 18 h at 28°C, the CFP signals were captured under a confocal laser scanning microscope at an excitation wavelength 405 nm (Leica TCS SP5, Wetzlar, Germany).

Super-resolution structured illumination microscopy and image analysis

Immunofluorescence assays were conducted by the procedures described above. Meiotic cells with high-quality immunofluorescence signals were marked for recording approximate X-Y positions using the ZEISS A2 microscope. Then these cells were repositioned under the DeltaVision microscope (OMXTM V4; GEhealthcare, Chicago, IL, USA) and the super-resolution images were processed with SoftWoRx (Applied Precision, Issaquah, WA, USA). And the distance between axial elements was measured using IMAGEJ 1.37a software.

RNA extraction and real-time PCR

Real-time PCR assays were used to detect the expression patterns of SCC3. Total RNA was extracted from different tissues in wild-type including root, stem, leaf, sheath, spikelet, 2 cm long panicle, 2-3 cm long panicle, 3-5 cm long panicle, 5-7 cm long panicle and 7 cm long panicle using TRIzol reagent. After digestion with RNase-free DNaseI to remove genomic DNA, reverse transcription PCR was performed to synthesize cDNA using SuperScriptTM III Reverse Transcriptase by manufacturer’s protocol (Invitrogen). Quantitative real-time PCR assays were performed on the Bio-Rad CFX96 instrument using Evagreen (Biotium, Freemont, CA, USA) with a procedure of 98 °C 3min, 40 cycles of 98 °C 15s and 60 °C 20s. The UBIQUITIN gene was used as an internal reference. Three biological repeats were performed for each sample. Primers used for real-time PCR assays were listed in Table S1.

Phylogenetic analysis

The full-length amino acid sequence of SCC3 was selected as a template and searched by the NCBI blastp tool to identify similar proteins among species. A detailed protocol for further analysis has been reported previously (Li et al., 2018b; Ren et al., 2018; Ren et al., 2020). Filtered sequences were downloaded and used for constructing neighbor-joining trees by MEGA5 software and processed with EvolvieW (https://evolgenius.info/evolview/).

Multiple sequence alignments

Multiple alignments were performed using the online toolkit MAFFT (https://toolkit.tuebingen.mpg.de/mafft) and processed with ESPRIPT3 (http://espript.ibcp.fr/ESPript/ESPript/) by previously detailed reported (Zhao et al., 2021).

Experimental model and subject details

The rice scc3 mutant was produced by CRISPR-Cas9 toolkit in japonica rice variety Yandao 8. Primers used for CRISPR-Cas9 are listed in Table S1. Other meiosis mutants pair3 (Wuxiangjing 9), pair2 (Nipponbare), zep1 (Nipponbare), rec8 (Yandao 8), sgo1 (Yandao 8), spo11-1 (Guangluai 4), dmc1 (Nipponbare) have been previously described (Wang et al., 2016). Yandao 8 was used as the wild-type. All the rice plants were grown in a paddy field under natural rice-growing conditions.

Quantification and statistical analysis

At least three independent biological repeats were performed for each experiment. Values in Figures are means ± SD. The significant difference between the wild type and corresponding mutants was evaluated by two-tailed Student’s t-test and p < 0.05 was considered as a criterion for judging statistically significant difference. The significance labeled in the Figures represents the following meanings: *, p < 0.05; **, p < 0.01; ***, p < 0.001; ****, p < 0.0001.

eLife Assessment

This important study elucidates the function of the cohesin subunit SCC3 in impeding DNA repair between inter-sister chromatids in rice. The observation of sterility in the SCC3 weak mutant prompted an investigation of abnormal chromosome behavior during anaphase I through karyotype analysis. While the evidence presented is largely solid, the strength of support can be substantially improved in some aspects, leaving room for further investigation. This research contributes to our understanding of meiosis in rice and attracts cell biologists, reproductive biologists, and plant geneticists.

Reviewer #1 (Public Review):

Summary:

The manuscript describes the identification and characterization of rice SCC3, including the generation and characterization of plants containing apparently lethal null mutations in SCC3 as well as mutant plants containing a c-terminal frame-shift mutation. The weak scc3 mutants showed both vegetative and reproductive defects. Specifically, mitotic chromosomes appeared to partially separate during prometaphase, while meiotic chromosomes were diffuse during early meiosis and showed alterations in sister chromatid cohesion, homologous chromosome pairing, and recombination. The authors suggest that SCC3 acts as a cohesin subunit in mitosis and meiosis, but also plays more functions other than just cohesion.

Strengths:

The manuscript contains a large amount of generally high-quality data.

Weaknesses:

Several of the conclusions drawn in the manuscript are not supported by the data. There are many examples where the authors either draw conclusions or make statements that are just not justified based on the data presented or present a conclusion as a new finding, which has already been demonstrated in the past by others. For example, they claim that SCC3 functions in the maintenance of replication. From my reading of the manuscript, nowhere did the authors examine DNA replication. Likewise, several of the conclusions drawn are in direct contrast with what is known about SCC3 in other organisms. Therefore, the conclusions are either groundbreaking or incorrect.

Reviewer #2 (Public Review):

Summary:

This manuscript shows detailed evidence of the role of cohesin regulators in rice meiosis and mitosis.

Strengths:

There is a very clear mechanism for its role during replication. The strength of the evidence and its novelty is very high. This paper makes a significant contribution to the body of knowledge on meiotic cohesion in a valuable plant model.

Weaknesses:

The authors did not consider creating heterozygous mutants for the replication fork.

Moderate English language editing may be required.

Reviewer #3 (Public Review):

Summary:

Prior research on SCC3, a cohesin subunit protein, in yeast and Arabidopsis has underscored its vital role in cell division. This study investigated into the specific functions of SCC3 in rice mitosis and meiosis. In a weakened SCC3 mutant, sister chromatids separating was observed in anaphase I, resulting in 24 univalents and subsequent sterility. The authors meticulously documented SCC3's loading and degradation dynamics on chromosomes, noting its impact on DNA replication. Despite the loss of homologous chromosome pairing and synapsis in the mutant, chromosomes retained double-strand breaks without fragmenting. Consequently, the authors inferred that in the scc3 mutant, DNA repair more frequently relies on sister chromatids as templates compared to the wild type.

Strengths:

The study presents exceptionally well-executed research in the field of rice cytogenetics.

Weaknesses:

While the paper's conclusions are generally well-supported, further substantiation is needed for the claim that SCC3 inhibits template choice for sister chromatids. To bolster this conclusion, I recommend that the authors perform whole-genome sequencing on parental and F1 individuals from two rice variants, subsequently calculating the allele frequencies at heterozygous sites in the F1 individuals. If SCC3 indeed inhibits inter-sister chromatid repair in the wild type, we would anticipate a higher frequency of inter-homologous chromosome repair (i.e., gene conversion). This should be manifested as a bias away from the Mendelian inheritance ratio (50:50) in the offspring of the wild type compared to the offspring of the scc3+/- mutant.

Author response:

The following is the authors’ response to the original reviews.

We express our sincere appreciation for your insightful comments and constructive suggestions. It is with great pleasure that we submit the revised version of our manuscript. Over the past months, we have meticulously considered all the invaluable feedback provided by the three anonymous reviewers, and endeavored to incorporate significant revisions accordingly. Furthermore, we have meticulously rephrased the results section in accordance with your guidance, aiming to bolster the rigor of our manuscript. The specific changes implemented in the revised manuscript are outlined below:

- Revised the title of the manuscript.

- Revised the description of early mitotic and meiotic chromosome structure in the scc3 mutant (Lines 167-274).

- Added the BiFC results illustrating the interaction between SCC3 and other cohesin proteins in Figure S10.

- Enhanced the detail in the description of figure legends, particularly for Figures 2 and 4.

- Refined and rephrased the language of the manuscript.

We hope these positive revisions have substantially strengthened the manuscript. Once again, we extend our heartfelt gratitude for your invaluable input.

eLife assessment

This important study elucidates the function of the cohesin subunit SCC3 in impeding DNA repair between inter-sister chromatids in rice. The observation of sterility in the SCC3 weak mutant prompted an investigation of abnormal chromosome behavior during anaphase I through karyotype analysis. While the evidence presented is largely solid, the strength of support can be substantially improved in some aspects, leaving room for further investigation. This research contributes to our understanding of meiosis in rice and attracts cell biologists, reproductive biologists, and plant geneticists.

Public Reviews:

Reviewer #1 (Public Review):

The manuscript describes the identification and characterization of rice SCC3, including the generation and characterization of plants containing apparently lethal null mutations in SCC3 as well as mutant plants containing a c-terminal frame-shift mutation. The weak scc3 mutants showed both vegetative and reproductive defects. Specifically, mitotic chromosomes appeared to partially separate during prometaphase, while meiotic chromosomes were diffuse during early meiosis and showed alterations in sister chromatid cohesion, homologous chromosome pairing, and recombination. The authors suggest that SCC3 acts as a cohesin subunit in mitosis and meiosis, but also plays more functions other than just cohesion.

Reviewer #2 (Public Review):

This manuscript shows detailed evidence of the role of cohesin regulators in rice meiosis and mitosis.

Reviewer #3 (Public Review):

Prior research on SCC3, a cohesin subunit protein, in yeast and Arabidopsis has underscored its vital role in cell division. This study investigated into the specific functions of SCC3 in rice mitosis and meiosis. In a weakened SCC3 mutant, sister chromatids separating was observed in anaphase I, resulting in 24 univalents and subsequent sterility. The authors meticulously documented SCC3's loading and degradation dynamics on chromosomes, noting its impact on DNA replication. Despite the loss of homologous chromosome pairing and synapsis in the mutant, chromosomes retained double-strand breaks without fragmenting. Consequently, the authors inferred that in the scc3 mutant, DNA repair more frequently relies on sister chromatids as templates compared to the wild type.

We extend our sincere gratitude to the Editors and the Reviewers for their highly constructive and insightful suggestions. We deeply appreciate receiving both positive feedback and constructive criticism on our manuscript. In light of the reviewers’ comments, we have diligently undertaken substantial revisions to improve the manuscript. The revised version comprehensively addresses all the points raised by the reviewers.

Below, we provide a detailed point-by-point response to the reviewers’ comments:

Recommendations for the authors:

Reviewer #1:

(1) Line 170- looking at pollen formation does not specifically evaluate whether SCC3 is involved in meiosis.

Thank you very much for this advice. We totally agree with your point of view that pollen formation defects only indicate the problem of gametogenesis. We are sorry for not accurately describing this sentence. It has been revised in the manuscript (Lines 167-176).

(2) Lines 203-205- this seems more like discussion and is pure speculation. Another possibility described above is that the truncated SCC3 protein is partially functional and what they see is due to this partial functionality. Have the authors considered the possibility that a partially functional version of SCC3 is produced that alters its function or the function of the cohesin complex? How much of the protein epitope remains in the truncated protein?

We are so grateful for the insightful suggestions provided. We concur with the proposition that a partially functional SCC3 may indeed be synthesized, contributing to the survivability of the mutant. Notably, the truncated version of the protein retains approximately 60% to 70% of the epitope, which ostensibly maintains a residual functionality within the weak scc3 mutant. In this manuscript, the loss of C-terminal 910-1116 aa of SCC3 contains a special protein epitope and a certain protein secondary structure, which may alter the protein’s folding and its subsequent roles within the cohesin complex.

In this study, we encountered challenges in generating null alleles of the scc3 mutants in rice utilizing the CRISPR-Cas9 system. Consequently, it is plausible that the scc3-1 and scc3-2 variants represent null alleles of SCC3, resulting in embryonic lethality. We posit that the identification of weak alleles is paramount to facilitating the survival of the organism. Thus, selecting some weak mutants, particularly those exhibiting the most pronounced phenotype, is advantageous for conducting further research. Our findings indicate that the diminished scc3 mutant lacks only a segment of the C-terminal, yet this deficiency is adequate to ensure the plant's survival while significantly impeding the meiotic process. We cannot dismiss the likelihood that these observed defects are attributable to the unique truncated proteins. We extend our sincerest thanks once again.

(3) Lines 212- I question whether what the authors see in Figure 2 is chromosome fragmentation. It could just as well be alterations in chromosome structure. Likewise, the authors provide little to no evidence that the mutation affects the replication process. Rather, the presence of replicated chromosomes later in mitosis and meiosis would argue that replication is not disrupted.

We express our gratitude to the reviewer for highlighting this critical inquiry. Contrary to the scenario of chromosome fragmentation, as you astutely observed, the preservation of normal sister chromatids during prometaphase indicates that the replication process remains uninterrupted. In alignment with your insights, our study embarked on an extensive series of full-length fluorescence in situ hybridization (FISH) experiments to elucidate the underlying mechanisms contributing to the observed increase in the distance between sister chromatids, particularly during interphase. The preponderance of our findings corroborates the hypothesis that the chromosomes exhibit alterations in structure, as depicted in Figure 2A. Intriguingly, our data suggest that cohesin, upon interaction with other chromatin-bound proteins, may facilitate loop extrusion, anchoring itself in a manner that potentially alters chromosomal architecture. These alterations in chromosome structure and the subsequent defects in genome folding and cohesion establishment, particularly rely on SCC3. In response to your valuable suggestions, we have meticulously revised the relevant sections of our manuscript. We extend our sincere thanks for your insightful comments.

(4) Line 230- what does the sentence SCC3 may enhance the interaction with DNA mean, the interaction of the cohesin complex?

We are sorry for the ambiguity in our initial description and wish to clarify that SCC3 indeed plays a pivotal role in augmenting the interaction between the cohesin complex and DNA. Our observations revealed an upsurge in the signal intensity of SCC3 as cells transition from interphase to prophase, as depicted in Figure 2B. This enhancement correlates with the observed defects in scc3 mutants during prophase, suggesting that SCC3’s functional significance is particularly pronounced at this stage of the cell cycle. We have revised our manuscript to reflect these insights more accurately, in accordance with your valuable suggestions. We express our sincere gratitude for your guidance.

(5) Oddly, and unexplainably the authors present data indicating that SCC3 interacts with RAD21.1, but not SMC1, SMC3, or REC8. The fact that the authors report that SCC3 only interacts with RAD21.1 but no other cohesin proteins is quite hard to explain.

As argued in the point above, the available data do not provide compelling evidence supporting the interaction between SCC3 and other cohesin proteins. We have repeated yeast two-hybrid (Y2H) experiments yielding consistent outcomes, which also surprised us initially. In the revised manuscript, we further added the bimolecular fluorescence complementation (BiFC) results between SCC3 and other cohesin proteins in rice protoplast (Figure S10). These supplementary data affirm that SCC3 predominantly interacts with RAD21.1, excluding interactions with other cohesin proteins. While the absence of such interactions is perplexing, our investigations have failed to detect any binding between SCC3 and other cohesin proteins.

A weak interaction between SCC3 and REC8 has been reported in Arabidopsis (Kuttig et al. bioRxiv https://doi.org/10.1101/2022.06.20.496767). We speculate that either these proteins do not interact or the yeast-hybrid assays may be inadequate for detecting their interaction, as several factors can impede interaction in a heterologous system. In Figure 7, we could only detect the interaction between SCC3 and RAD21.1 in both Y2H and BiFC experiments. This suggests potential alterations in protein folding or conformation, or the involvement of additional regulatory factors modulating the interaction between SCC3 and other cohesin proteins. Notably, given RAD21.1’s pivotal role as a core component in the cohesin complex, our supplementary findings demonstrate the interactions between SMC1/3 and RAD21.1 (data not shown). Consequently, our current data propose a model wherein RAD21.1 and SMC1/3 form a circular structure, with SCC3 positioned on the outer periphery of the ring complex, associating specifically with RAD21.1 (Figure 8A).

Reviewer #2:

The authors did not consider creating heterozygous mutants for the replication fork. Moderate English language editing may be required.

We extend our gratitude to the reviewer for their valuable suggestions. Initially, we did not explore the potential relationship between SCC3 and the replication fork. Cohesin, as we understand, becomes associated with DNA prior to DNA replication. The phenomenon of sister chromatid co-entrapment arises as replication forks traverse through cohesin rings, a process intricately linked to DNA replication dynamics. In this study, we exclusively observed aberrant chromosome structures in the scc3 mutant during interphase (Figure 2). We conjecture that these anomalies may stem from alterations in chromosome structure, such as genome folding and loop extrusion, rather than being directly attributable to the DNA replication fork. However, the precise nature of these chromosome structural aberrations during interphase in the scc3 mutant remains elusive, necessitating further comprehensive investigation in future studies. We have refined the language of our manuscript in accordance with the reviewer’s suggestions. Once again, we express our sincere appreciation for the invaluable suggestions provided.

Reviewer #3:

While the paper's conclusions are generally well-supported, further substantiation is needed for the claim that SCC3 inhibits template choice for sister chromatids. To bolster this conclusion, I recommend that the authors perform whole-genome sequencing on parental and F1 individuals from two rice variants, subsequently calculating the allele frequencies at heterozygous sites in the F1 individuals. If SCC3 indeed inhibits inter-sister chromatid repair in the wild type, we would anticipate a higher frequency of inter-homologous chromosome repair (i.e., gene conversion). This should be manifested as a bias away from the Mendelian inheritance ratio (50:50) in the offspring of the wild type compared to the offspring of the scc3+/- mutant.

We express our sincere appreciation for your insightful suggestions. It is really a good suggestion. We have arranged to do this experiment. As it takes long time to prepare plant materials and sequence analysis, we hope the ongoing sequencing work will get some important information supporting those hypotheses. As we have not obtained the direct evidence that SCC3 involved in sister chromatid repair, we changed the title as “SCC3 is an axial element essential for homologous chromosome pairing and synapsis”. Once again, we really extend our gratitude for your invaluable suggestions.

A point that warrants consideration is the placement of the protein interaction experiments involving SCC3 within the paper. It is presented relatively late in the manuscript. If the authors possess information regarding the interaction between RAD21.1 and SCC3 and how it relates to the functional study of RAD21.1, it could contribute to a more comprehensive analysis. However, if this information is unrelated to the current study, it might be advisable to omit it, as it appears to diverge from the main focus of this work.

We express our sincere gratitude for your invaluable suggestions. It has been documented in yeast that the interaction between SCC3 and SCC1 is indispensable for the efficient loading of cohesin. In our study, we endeavored to elucidate the intricate relationships among various cohesin subunits. Through our investigations, we have discerned that RAD21.1 serves as a pivotal core subunit within the cohesin complex, facilitating interactions with both SMC1/3 and SCC3 (data not shown). Additionally, our findings indicate that the interaction between RAD21.1 and SCC3 is imperative for maintaining the stability of the cohesin ring and its association with DNA (data not shown). Consequently, the interaction between these two proteins assumes paramount importance for our subsequent analyses. This study holds significant promise for future investigations.

It's worth noting that while the title of the study claims that "SCC3 inhibits inter-sister chromatids repair during rice meiosis," the last sentence of the abstract weakens this conclusion by using the word "seems." A study's title should ideally reflect the most definitive and conclusive findings.

We sincerely appreciate your valuable suggestions. In response, we have revised the description in our manuscript to enhance its rigor.

In Figure 8C, it appears that cohesin is depicted between two DNA strands.

Figure 8C illustrates the process of sister chromatid repair during meiosis in the scc3 mutant. Two gray lines and two blue lines represent the four sister chromatids of two homologous chromosomes, respectively. In the wild type, cohesin plays a crucial role in tethering together the two sister chromatids. As per your reminder, cohesin should indeed encircle the two sister chromatids, as depicted in Figure 8B. Following a thorough evaluation and to mitigate any potential confusion, we have deleted Figure 8C.

Version 2

Abstract

Abstract

Cohesin is a multi-subunit protein that plays a pivotal role in holding sister chromatids together during cell division. Sister chromatid cohesion 3 (SCC3), constituents of cohesin complex, is highly conserved from yeast to mammals. Since the deletion of individual cohesin subunit always causes lethality, it is difficult to dissect its biological function in both mitosis and meiosis. Here, we obtained scc3 weak mutants using CRISPR-Cas9 system to explore its function during rice mitosis and meiosis. The scc3 weak mutants displayed obvious vegetative defects and complete sterility, underscoring the essential roles of SCC3 in both mitosis and meiosis. SCC3 is localized on chromatin from interphase to prometaphase in mitosis. However, in meiosis, SCC3 acts as an axial element during early prophase I and subsequently situates onto centromeric regions following the disassembly of the synaptonemal complex. The loading of SCC3 onto meiotic chromosomes depends on REC8. scc3 shows severe defects in homologous pairing and synapsis. Consequently, SCC3 functions as an axial element that is essential for maintaining homologous chromosome pairing and synapsis during meiosis.

Introduction

The correct segregation of sister chromatids is essential for mitotic cell stabilization and to generate haploid gametes during meiosis. The cohesin complex is a conserved multi-subunit protein system that guarantees the proper segregation of sister chromatids in mitosis across living organisms. This complex is formed by four core subunits, including the structural maintenance of chromosomes (SMC) subunits SMC1 and SMC3, and the sister chromatid cohesin (SCC) subunits SCC1 and SCC3 (Bolanos-Villegas et al., 2017; Moronta-Gines et al., 2019). The cohesin complex forms a ring encircling the sister chromatids and ensures genomic stability during DNA replication (Higashi et al., 2020). In addition, cohesin also participates in the repair of DNA double-strand breaks (DSB), homologous pairing, construction of the synaptonemal complex (SC), orientation of kinetochores, and regulation of gene expression (Nasmyth and Haering, 2009).

Cohesin is established during the S phase, and includes arm and centromeric cohesin. Arm cohesin ensures sister chromatids bond together during mitotic prophase. Centromeric cohesin persists until the metaphase-to-anaphase transition, and is vital to resist the force exerted by spindle microtubules, allowing for the accurate segregation of sister chromatids to opposite poles of the cell. The dissociation of cohesin generally occurs in two steps, including through the dissociation of cohesin on chromosome arms during prometaphase, followed by the degradation of cohesin at the centromeres at the onset of anaphase (Ishiguro and Watanabe, 2007; Watanabe, 2005). This allows sister chromatids to segregate in mitosis.

Meiosis contains a single round of DNA replication, but, in contrast to mitosis, is followed by two successive cell divisions. Nonetheless, while the loading mechanism of cohesin is very similar to the analogous process during mitosis, the cohesin degradation during meiosis occurs in two steps: from the chromosome arms during the first meiotic division, and from the centromeres during the second meiotic division (Mercier et al., 2015). During meiosis I, homologous chromosomes are linked by chiasmata that separate at metaphase I. At this stage, cohesin dissociates from chromosome arms while the two sister chromatids remain closely together because centromeric cohesin is preserved. Subsequently, the sister chromatids separate at anaphase II due to the complete cohesin disassociation at both chromosome arms and centromeres. During this process, sister chromatid cohesin is protected by shugoshin from cleavage by separases (Wang et al., 2012; Wang et al., 2011), which ensures the correct separation of sister chromatids during meiosis.

Notably, the cohesin complex plays additional functional roles in meiosis. Specifically, cohesin on chromosome arms is crucial for normal progression of meiosis I, while centromeric cohesin plays a more prominent role in meiosis II by geometrically orientating the sister chromatids. During meiosis I, homologous chromosomes are segregated in opposite directions to halve the chromosome number, but sister chromatids remain temporarily mono-oriented and move to the same pole. In meiosis II, sister chromatid pairs become bi-oriented and separate towards opposite spindle poles through similar mechanisms as those observed in mitosis (Schvarzstein et al., 2010). In this process, centromeric cohesin plays a crucial role in establishing kinetochore geometry and thus determining sister kinetochore orientation (Gryaznova et al., 2021). In addition, the cohesin complex affects the formation of chromosome axial elements (AE) and the assembly of synaptonemal complex (SC) (Agostinho et al., 2016; Fujiwara et al., 2020; Llano et al., 2012). The cohesin complex is now emerging as a key player in mediating homolog bias of meiotic recombination (Hong et al., 2013; Phipps and Dubrana, 2022; Sanchez-Moran et al., 2007). Besides, the cohesin complex regulates gene expression in eukaryotes, acting as a transcriptional co-activator in humans that is recruited by sequence-specific transcription factors (Casa et al., 2020; Kagey et al., 2010; Lara-Pezzi et al., 2004; Remeseiro et al., 2012).

To date, the relationship between discrete cohesin subunits and their role in promoting sister chromatid cohesion remains elusive. In yeast, SCC1 ensures sister chromatid cohesin in mitosis and associated with chromosomes during the S phase, dissociating during the metaphase-to-anaphase transition. The expression of SCC1 gradually decreases at the beginning of meiosis, suggesting a reduced role for SCC1 during meiosis (Michaelis et al., 1997). Instead, REC8, a specialized meiotic-specific cohesin component, replaces SCC1 in yeast (Klein et al., 1999; Watanabe and Nurse, 1999). REC8 also plays a crucial role establishing sister kinetochore orientation in yeast, mice and plants (Chelysheva et al., 2005; Ogushi et al., 2021; Toth et al., 2000). Additionally, rec8 mutants in flowering plants exhibit sticky chromosomes, which severely affects the localization of other key meiotic proteins (Cai et al., 2003; Chelysheva et al., 2005; Shao et al., 2011). SCC3 acts as an important component of the cohesin complex, which has been rarely studied in plants. The only research in Arabidopsis shows that SCC3 is responsible for maintaining centromeric orientation during meiosis (Chelysheva et al., 2005). However, the mechanisms through which SCC3 regulates the assembly of sister chromatid cohesion and its function in meiosis remain poorly understood. Whether SCC3 is similar to REC8 in regulating meiotic chromosome behavior in flowering plants remains unclear.

Here, we investigated the functional roles of SCC3 in rice, elucidating its participation in both mitosis and meiosis. Through our examination of chromosome 12 replication during interphase, we confirmed SCC3’s role in stabilizing the chromosome structures, which is pivotal for sister chromatid cohesion in both mitosis and meiosis. Additionally, we explored the dynamics of cohesin loading and degradation on chromosomes using the antibody of SCC3 during mitosis and meiosis. SCC3 emerges as an axial element intricately involved in homologous pairing, synapsis, recombination progress, and crossover (CO) formation. Our findings also shed light on the regulation of SCC3 localization by REC8. Meanwhile, our investigation revealed the efficient repair of meiotic DSBs in scc3, utilizing sister chromatids as repair templates. We propose that SCC3 acts as a constituent of the cohesin complex, which plays an indispensable role in meiotic homologous pairing and synapsis.

Results

SCC3 causes both vegetative and reproductive growth defects

Considering SCC3 is a highly conserved protein in different species, we performed a BLAST search in the Rice Genome Annotation Project database (http://rice.plantbiology.msu.edu/cgibin/gbrowse/rice) and found a single hypothetical homologue of SCC3 proteins (A. thaliana SCC3, M. musculus STAG3 and H. sapiens SA1). This candidate protein is encoded by the Os05g0188500 gene and shares the highest similarity with AtSCC3. The full-length cDNA of SCC3 is 3755 nucleotides long and comprises 22 exons and 21 introns. SCC3 contains a 3345-nucleotide-long ORF that encodes a 1116 amino-acid protein. In higher eukaryotes, SCC3 is an evolutionary conserved protein that contains a highly conserved STAG domain (140-239aa), as predicted by SMART (Figure S1A). Multiple alignments of full-length SCC3 protein sequences and its homologs in other plants revealed the STAG domain is highly conserved across both monocotyledons and dicotyledons (Figure S1B). Homology modeling indicated that the structure of the SCC3 protein is strongly conserved among different species, which also possess similar STAG domains (Figure S2A). Additionally, a phylogenetic tree was constructed with full-length SCC3 amino-acid confirming that confirmed this protein is conserved in eukaryotes (Figure S2B). We also performed multiple sequence alignments of these proteins used in the evolutionary tree analysis (Figure S3).

A previous study in Arabidopsis (Chelysheva et al., 2005) identified a T-DNA insertion at the boundary between intron 5 and exon 6 of SCC3 that caused lethality, whilst a weak allele was able to survive albeit with some developmental defects. To clarify the function of SCC3 in rice, we used the CRISPR-Cas9 system to generate three transgenic lines (Figure S4). The first allele (scc3-1) contains a frameshift “GA” deletion in exon 2 and the second allele (scc3-2) contains a frameshift “ACCGA” deletion in exon 11. We tested for allelism between these two scc3 mutants by crossing scc3-1+/- (male) and scc3-2+/- (female). Of the 78 F1 plants, 18 were scc3-1+/-, 26 were scc3-2+/-, 34 were wild-type for both loci. Thus, these two mutants are allelic and the heterozygous scc3-1/scc3-2 is lethal. We were unable to isolate homozygous mutants in these two allelic progenies (n=128, 87 plants were scc3-1+/- and 41 were wild-type; n=116, 80 plants were scc3-2+/- and 36 were wild-type). These results suggest that scc3-1 and scc3-2 mutants are both embryo-lethal.

In addition, we obtained a weak transgenic line (scc3) containing a frameshift “T” insertion in exon 19 leading to premature termination of translation (Figure S4). We found the scc3 weak mutant exhibits abnormal vegetative growth (Figures 1A and 1B), and proceeded with further investigation of its other phenotypic effects. This revealed that scc3 decreases plant height, tiller number and panicle length (Figures 1C and S5B-S5D). To investigate the patterns of SCC3 expression, we employed RT-PCR in various tissues and found the gene is ubiquitously expressed, particularly in roots, leaves and panicles (Figure S5A). Accordingly, the weak mutant of SCC3 causes severe vegetative growth defects in rice plants.

SCC3 is required for sister chromatid cohesion during mitosis

(A) Morphology of seedlings and root tips in wild type and scc3.

(B) Morphology of plants in wild type and scc3.

(C) Morphology of panicles in wild type and scc3.

(D) Pollen grains stained in 1% I2-KI solution in wild type and scc3. Bars, 50 μm.

(E) The chromosome behaviors of root tip cells in wild type and scc3, stained with DAPI. Individual chromosomes are marked by red boxes. The distance between the sister chromatids increased significantly at prometaphase, and the sister chromatids separated in advance at metaphase in scc3. Bars, 5 μm.

(F) FISH analysis of mitotic cells with centromere-specific probes in the wild type and scc3. The distance between centromeres increased significantly in scc3. CentO (red) signals indicate centromeres. Chromosomes were stained with DAPI (blue). Bars, 5μm.

(G) The distance between sister chromatids in different mitotic chromosomes. In normal condition, two sister chromatids form tight stick. In scc3, the distance between chromosome arms and telomeres were increased significantly. Curve diagrams show the distance between the arms and the telomeres, respectively, as measured by IMAGEJ. Bars, 1 μm.

(H) Graphical representation of the frequency of each type of chromosome morphology. The classification was assigned when >50% chromosomes in a spread showing the indicated morphology.

To elucidate whether gametogenesis is normal in scc3, we performed cytological observation of anthers stained with 1% I2-KI, and found almost all pollen grains were shrunken and inviable (Figures 1D and S5E). Our results also showed scc3 flowers did not set seeds when pollinated with wild-type pollen, suggesting the mutant was both male and female sterile. We further observed heterozygous scc3 plants produced progenies with a segregation ratio of 3:1 (normal: dwarf and sterile), indicating the scc3 mutation is recessive and monogenic. These data suggest that the scc3 weak mutation dramatically interferes with both plant vegetative and reproductive growth.

SCC3 is required for sister chromatid cohesion during mitosis

To determine whether SCC3 is involved in sister chromatid cohesion during mitosis, we observed chromosome behavior in root tip cells of scc3 and wild-type plants. In wild-type plants, DNA replicates and folds into an ordered structure during interphase (Figure 1E). Subsequently, chromatin condense to form rod-like chromosomes during prophase, further align at the equatorial plate during metaphase before segregating to distinct poles of the cell.

In contrast, scc3 exhibited “X”-shaped chromosomes with distant arms and telomeres at prometaphase (Figures 1E and 1G), indicating compromised cohesion between sister chromatids. Notably, approximately 70.6% of scc3 root tip cells (n=88) exhibited partially or completely separated sister chromatids (Figure 1H). During the transition from prometaphase to metaphase, all sister chromatids prematurely separated and failed to align properly at the equatorial plate. From anaphase to telophase, sister chromatids were loosely pulled towards the spindle poles (Figure 1E). However, aneuploidy was not observed after anaphase, as mitotic tubulin remained attached to the centromeres of sister chromatids at metaphase, ensuring their equal segregation.

Additionally, we monitored the distance between centromeres using FISH probes with the centromere-specific tandem repetitive sequence CentO in both wild type and scc3 somatic cells during prometaphase. We observed a substantial increase in the distance between sister centromeres in scc3 (Figure 1F, n=25). However, despite the distance between sister chromatids increased in scc3, sisters were still observed side-by-side during prometaphase, suggesting the presence of other cohesion proteins maintaining sister chromatid association. Our results demonstrated SCC3 is required for cohesion between sister arms and centromeres during mitosis.

To elucidate the underlying cause of the increased distance between sister chromatids, we performed full-length FISH assays to monitor the dynamic phenotype of chromosome 12 during interphase. In the wild type, replicating sister chromatids were observable during interphase (Figure 2A). Subsequently, two chromosome clumps formed during prophase and developed into high-order linear chromosomes during prometaphase. During this period, the two sister chromatids overlapped and condensed into a short stick. However, in scc3, no chromosome clumps were detected during interphase, and an evident uncondensed chromosome structure was observed, suggesting a role for SCC3 in organizing genome topology and chromatin association dynamics. During prophase, chromosome 12 remained in a loose lump and was unable to condense into a stick-like structure. At prometaphase, the abnormal chromatin finally formed loosely-connected sister chromatids (Figure 2A). These observations implicate SCC3 involving in the dynamic structural change of chromatin during interphase, influencing the ability of cohesin to bind sister chromatids, potentially regulating chromatin composition and genome dynamics.

SCC3 alters the structure of sister chromatids during mitosis

(A) The dynamic process of chromosome structure in early mitosis as revealed by pooled oligos specific to chromosome 12 (red). In the wild type, S1 and S2 indicate the replicated sister chromatids. However, in scc3 sister chromatids exhibited the variation of structure from interphase to prophase. Mitotic chromosomes in wild type and scc3 were stained with DAPI (blue). Bars, 5 μm.

(B) The loading pattern of SCC3 (green, from mouse) and CENH3 (red, from rabbit) in wild-type root tip cells. Chromosomes were stained with DAPI (blue). Bars, 5 μm.

To investigate the localization of SCC3 during mitosis, we generated a mouse polyclonal antibody against the C-terminal of SCC3 and performed immunostaining in wild-type root tip cells (Figure 2B). No SCC3 signal was detected in scc3 root tip cells, confirming the specificity of the SCC3 antibody utilized in mitosis (Figure S6A). In most instances, SCC3 was uniformly distributed within the nuclear during interphase. As cell division started, SCC3 immunosignals around centromeric regions intensified at prophase. SCC3 eventually coalesced into intact chromosomes at prometaphase. From metaphase to telophase, SCC3 signals completely disappeared (Figure 2B). These observations suggest that SCC3 may enhance the cohesin-DNA interaction during the transition from interphase to prophase and gradually dissociates from chromosomes after prometaphase.

SCC3 acts as an axial element during meiosis

To evaluate SCC3 localization during meiosis, we performed dual-immunolocalization between SCC3 and REC8 in wild-type meiocytes. REC8 is a specific axial protein specific to meiosis (Shao et al., 2011). REC8 signals were initially detected as foci in the nucleus at leptotene, during which SCC3 was simultaneously detectable at the same time (Figure 3A). From leptotene to zygotene, SCC3 signals were detected as similar elongated foci that formed filamentous structures on the chromosomes. During pachytene and diplotene, constant SCC3 signals were distributed along the chromosome axes (Figure 3A). At diakinesis, axial SCC3 signals became diffuse and centromeric signals gradually became clear. At metaphase I, only centromeric SCC3 signals remained (Figure 3B). However, REC8 signals were difficult to be detected after diakinesis (Figure S7). In order to further clarify the localization pattern between SCC3 and REC8, we performed super-resolution imaging on wild-type meiocytes. We found REC8 and SCC3 signals were parallel and colocalized forming the parallel axis along chromosome length (Figure 3C). In addition, the central element ZEP1 formed two parallel lines surrounded by SCC3 signals (Figure 3D). No SCC3 signal was present in scc3 meiocytes, confirming the specificity of the SCC3 antibody utilized in immunodetection (Figure S6B).

SCC3 is an axial element during meiosis

(A) In meiosis, SCC3 (green, from mouse) colocalizes with REC8 (red, from rabbit) from the leptotene to diplotene. Bars, 5 μm.

(B) During diakinesis and metaphase I, SCC3 (green, from mouse) gradually dispersed and finally retained in the centromeres indicated by CENH3 (red, from rabbit). Chromosomes were stained with DAPI (blue). Bars, 5 μm.

(C) Immunostaining of SCC3 (cyan, from mouse) and REC8 (red, from rabbit) in wild-type meiocytes at pachytene. Two parallel linear SCC3 signals colocalize with the REC8 linear signals, indicating chromosomal axial elements. Magnified images of the blocked regions are shown on the right. Bars, 5 μm.

(D) Immunostaining of SCC3 (cyan, from mouse) and ZEP1 (red, from rabbit) in wild-type meiocytes at pachytene. Two linear SCC3 signals wrap the ZEP1 signals, indicating central elements of SC are wrapped by axial elements. Magnified images of the blocked regions are shown on the right. Bars, 5 μm.

SCC3 is required for sister chromatids cohesion in early meiosis

The premeiotic interphase bears the resemblance to mitosis, wherein sister chromatids undergo replication and are tethered together by cohesins. To determine whether SCC3 is involved in sister chromatid cohesion during meiosis, we scrutinized the chromosome dynamics of meiocytes in both scc3 and wild type before leptotene. Detailed cytological phenotypes of meiotic initiation in rice have been previously elucidated (Zhao et al., 2018). In the wild type, the initial discernible stage of meiosis is preleptotene, during which sister chromatids have completed their replication. At this stage, each chromosome exhibited a condensed configuration, similar to the mitotic prophase chromosome (Figure 4A). Subsequently, they turned into fine threads at leptotene. However, in scc3, chromosomes showed defused morphology without clear outlines at preleptotene. These observations suggest that SCC3 affects the morphological construction of early meiotic chromosomes.

SCC3 is required for sister chromatid cohesion in early meiosis

(A) Meiocyte chromosome morphology at the preleptotene and leptotene in wild type, scc3, rec8 and rec8 scc3. Chromosomes at preleptotene exhibited a hairy rodlike appearance and formed thin threads at leptotene in wild type. However, chromosomes in scc3 were dispersed with blurred outlines and failed to from thin threads at leptotene. Bars, 5 μm.

(B) The dynamic process of chromosome structure in preleptotene was revealed by pooled oligos specific to chromosome 12 (red). Preleptotene chromosomes in wild type, scc3, rec8 and rec8 scc3. Chromosomes were stained with DAPI (blue). Bars, 5 μm.

We also scrutinized the chromosome behavior in both the rec8 single mutant and the scc3 rec8 double mutant at preleptotene (Figure 4A). Surprisingly, rec8 exhibited elongated chromosomes similar to the wild type, whereas scc3 rec8 displayed diffused chromosomes same as scc3. These results suggest that REC8 does not impede the early replication process of sister chromatids, and the chromosomal abnormalities observed in the scc3 rec8 double mutant would be attributed to the SCC3 mutation. We further performed full-length FISH assays to monitor chromosome 12 at preleptotene (Figure 4B). The labeled chromosome 12 formed short rod-shaped threads in both wild type and rec8, whereas it appeared diffuse structure in both scc3 and scc3 rec8. These data together with those from DAPI staining, further prove that SCC3 has an important function on chromosome structure during early meiosis, which is independent of REC8.

Homologous chromosome pairing and synapsis are disturbed in scc3

To elucidate the causes of sterility in scc3, we investigated the male meiotic chromosome behavior of wild-type and scc3 meiocytes. In wild-type meiocytes, homologous chromosomes initiated pairing during the early zygotene stage and completed synapsis at pachytene. Additionally, at metaphase I, 12 bivalents were methodically arranged on the equatorial plate, facilitating the segregation of homologous chromosomes during anaphase I (Figure 5A). The chromosomal dynamics in scc3 did not deviate from the wild type during zygotene. However, the onset of meiotic anomalies became evident at pachytene, characterized by the dispersion of most chromosomes as solitary filaments due to inadequate homologous pairing (Figure 5A). Twenty-four univalents were observed from diakinesis to metaphase I (n=161, 150 meiocytes had 24 univalents, while 11 meiocytes had 1-2 bivalents). Two sister chromatids of each univalent were always pulled to opposite poles leading to 48 sister chromatids being observed at anaphase I. Notably, in spo11-1 mutants implicated in the disruption of meiotic DSB formation and similarly exhibiting 24 univalents as observed in scc3, the separation of sister chromatids occurred either haphazardly or not at all (Figure S8A). These findings indicate that the mutation of SCC3 will lead to premature separation of sister chromatids at anaphase I.

Homologous pairing and synapsis are disturbed in scc3

(A) Meiotic chromosome behavior in wild type and scc3. Bars, 5 μm.

(B) Immunolocalization of ZEP1 (green, from mouse) and REC8 (red, from rabbit) in wild type and scc3 meiocytes. ZEP1 was severely suppressed from early zygotene to pachytene in scc3. Bars, 5 μm.

The formation of the telomere bouquet constitutes an essential precondition for homologous chromosome pairing during early zygotene (Zhang et al., 2020). To ascertain the occurrence of bouquet formation, we employed fluorescence in situ hybridization (FISH) assays targeting telomeres with a specific probe (pAtT4) in both wild type and scc3. In the wild type, telomeres congregated to a specified region, culminating in the formation of a quintessential bouquet at the nuclear envelope during zygotene. Notably, scc3 mutants also demonstrated typical telomere bouquet formation (Figure S8B). Subsequently, we utilized 5S ribosomal DNA (5S rDNA) as a marker to evaluate homologous pairing. One overlapping 5S rDNA focus was observed in wild-type pachytene, but two separate foci were observed in scc3, indicative of unpaired homologous chromosomes (n=52; Figure S8C). This observation corroborates the function of SCC3 in facilitating homologous chromosome pairing.

Homologous chromosomes align and subsequently juxtaposed via a proteinaceous structure known as the synaptonemal complex (SC), which comprises two parallel arrays of axial elements (AE), termed lateral elements (LE) prior to synapsis. These elements lay the groundwork for SC assembly (Ren et al., 2021). Given SCC3’s identification as a component of the AE, we examined the SC assembly process by immunostaining with an antibody against ZEP1 in both scc3 and wild-type meiocytes. In the wild type, ZEP1 progressively elongated, forming brief linear signals during zygotene that fully extended along the paired homologous chromosomes by pachytene (Figure 5B). Conversely, scc3 mutants exhibited ZEP1 as isolated punctate foci at early zygotene, a pattern that persisted into pachytene (Figure 5B), underscoring the critical role of SCC3 in SC assembly. Furthermore, the localization of other AE components (PAIR2 and PAIR3) was also investigated, and these proteins displayed normal behavior in scc3 (Figure S9).

SCC3 is essential for recombination progress and CO formation

Since homologous pairing was impaired in scc3 meiocytes, we wondered whether recombination was normal in scc3. γH2AX is a marker that is used as a proxy to assess DSB formation. Immunostaining results revealed no significant difference in the quantity of γH2AX foci between the wild type and scc3 (265.4 ± 8.6, n=31 for scc3 meiocytes; 262.2 ± 7.2, n=20 for wild-type meiocytes) (Figure 6A). Furthermore, we examined the DSB resection factor (COM1), meiotic strand-invasion and exchange factors (RAD51, DMC1) and the interference-sensitive crossover (CO) marker (ZIP4) in both wild type and scc3. Immunostaining results revealed that the number of COM1 foci (290.3 ± 8.8, n=26 for scc3 meiocytes; 293.8 ± 9.4, n=23 for wild-type meiocytes) showed no significant difference between the wild type and scc3 (Figures 6A and 6B). However, a significant reduction in the foci of DMC1 (65.4 ± 17.1, n=25 for scc3 meiocytes; 281.2 ± 8.3, n=20 for wild-type meiocytes), RAD51 (55 ± 21.7, n=26 for scc3 meiocytes; 306 ± 11.9; n=24 for wild-type meiocytes) and ZIP4 (58 ± 17.9, n=23 for scc3 meiocytes; 316.6 ± 11.1, n=21 for wild-type meiocytes) was observed in scc3, suggesting a disruption in recombination progression and CO formation. Previous studies have identified HEI10 as a marker of Class I COs. In the wild type, prominent HEI10 foci were observed (24.2 ± 3.8, n=25) at late pachytene (Figures 6A and 6B). Conversely, scc3 predominantly exhibited sparse HEI10 foci, lacking in prominent, bright signals. Collectively, these results suggest that CO maturation was disturbed in scc3. Although a minor portion of recombinant proteins could be localized, the homologous recombination process was still greatly restricted.

SCC3 affects recombination progress and CO formation

(A) Immunolocalization of histone γH2AX (red), COM1 (green, from mouse), DMC1 (green, from mouse), RAD51 (green, from rabbit), ZIP4 (green, from mouse) and HEI10 (green, from mouse) in wild type and scc3. REC8 (red, from mouse and rabbit) signals were used to visualize chromosome axes. Bars, 5 μm.

(B) Box scatter plot of histone γH2AX phosphorylation, COM1, DMC1, RAD51 and ZIP4 in wild type and scc3. No difference of histone γH2AX and COM1 were shown between the wild type and scc3. DMC1, RAD51 and ZIP4 foci were significantly decreased in scc3 compared with wild type. Values are means ± SD. *** represents P < 0.001, two-tailed Student’s t-tests was performed.

(C) Chromosome behaviors in scc3, rec8, rec8 scc3, dmc1 and dmc1 scc3 from diakinesis to anaphase I. Bars, 5 μm.

Meiotic DSBs are repaired using sister chromatids in scc3

It is noteworthy that the formation of DSBs appears normal in scc3 (Figure 6A). However, the destabilization of homologous pairing and synapsis in scc3 leads to aberrations in homologous recombination. Despite the absence of the homologous recombination process, 24 univalents were observed from diakinesis to metaphase I. This was coupled with the complete separation of sister chromatids during anaphase I without chromosome fragmentation, suggesting proficient repair of these meiotic DSBs in scc3. These observations indicate that the meiotic DSBs in scc3 most likely utilize sister chromatid as repair templates. Additionally, we examined the chromosome behavior of both rec8 and rec8 scc3 from diakinesis to anaphase I (Figure 6C). They all exhibited abnormal chromosome morphology, manifesting a highly adhesive phenotype from diakinesis to metaphase I. Furthermore, chromosome fragmentation was observed in anaphase I, indicating inadequate repair of meiotic DSBs in both rec8 and rec8 scc3. Moreover, DMC1 has been implicated in inter-sister exchange during meiosis in both Arabidopsis and rice (Kurzbauer et al., 2012; Wang et al., 2016). To gain a deeper understanding of the relationship between DMC1 and SCC3 in rice, we generated the dmc1 scc3 double mutant and monitored the meiosis process from diakinesis to anaphase I. The meiotic defects of dmc1 and dmc1 scc3 were similar from diakinesis to metaphase I, resulting in 24 univalents (Figure 6C). However, the sister chromatids of dmc1 scc3 were completely separated in anaphase I, which was consistent with the phenotype of scc3. This suggests that the two genes might function independently in sister-chromatids repair during meiosis.

The loading of SCC3 onto meiotic chromosomes depends on REC8

After establishing SCC3’s role as an axial element in meiosis, an assessment was conducted to ascertain whether its localization was influenced by other meiotic AEs. REC8, PAIR2 and PAIR3 are known to localize on chromosome axes in rice, playing pivotal roles in facilitating SC assembly. The distribution of SCC3 signals remained unaffected in pair2 and pair3 mutants during meiosis (Figure 7B). Moreover, the distribution of SCC3 was normal in other mutants that affect meiotic DSB formation (spo11-1), strand invasion and homology searching (com1 and dmc1), and synaptonemal complex formation (zep1) (Figure 7B). This indicates that SCC3 localization is independent of the recombination process. However, a notable absence of SCC3 signal distribution was observed in rec8, indicating that REC8 plays a crucial role in the normal loading progression of SCC3 (n=36). Immunostaining of γH2AX confirmed normal DSB formation in rec8 (Figure 7A). In addition, the localization of the DSB end resection factor COM1 (n=25) and the inter-homologous strand-invasion factor RAD51 (n=26) was impaired in rec8 (Figure 7A). Furthermore, PAIR2 (n=28) only form short, interrupted stretches, and few detectable PAIR3 signals were observed (n=32). These findings indicate that meiotic DSB formation remains unaffected in rec8, while the processes of DSB end resection and homologous recombination are significantly compromised.

SCC3 loading onto meiotic chromosomes depends on REC8

(A) Immunolocalization of γH2AX (red, from rabbit), COM1 (green, from mouse), RAD51 (green, from rabbit), SCC3 (green, from mouse), PAIR2 (green, from rabbit) and PAIR3 (green, from rabbit) in rec8 meiocytes at zygotene. CENH3 (red and green, from rabbit and mouse) was used to indicate the centromeres. Bars, 5 μm.

(B) Immunolocalization of REC8 (red, from rabbit) and SCC3 (green, from mouse) in spo11, com1, dmc1, zep1, pair2 and pair3 meiocytes at zygotene. Bars, 5 μm.

(C) SCC3 interacts with RAD21.1 in yeast-two-hybrid assays. Interactions between bait and prey were examined on SD/DDO (SD-Leu-Trp) control media and SD/QDO (SD-Ade-His-Leu-Trp) selective media. AD, prey vector; BD, bait vector.

(D) Bimolecular fluorescence complementation assays show the interactions between SCC3 and RAD21.1 in rice protoplasts. Bars, 5 μm.

SCC3 interacts with SCC1

The cohesin complex is a ring-shaped protein complex where subunits interact with each other (Li et al., 2018a; Murayama and Uhlmann, 2015; Roig et al., 2014). To clarify the relationship between different subunits in rice, we performed Y2H assays between core cohesin proteins. Surprisingly, we found SCC3 only interacts with SCC1 (Figure 7C), which was further confirmed by BiFC assays (Figure 7D). However, it is worth noting that SCC3 could not interact with other cohesin proteins, including REC8, SMC1 and SMC3 (Figure 7C and Figure S10), indicating that SCC3 could function directly in combination with SCC1 in the cohesin complex.

Previous studies have shown that cohesins regulate gene expression (Cuadrado et al., 2015; Gligoris and Lowe, 2016; Remeseiro et al., 2012). Therefore, we investigated the expression levels of cohesin-related genes in wild-type and scc3 spikelets using RT-qPCR. Interestingly, we found that the cohesin subunits SCC1 and REC8, as well as the centromeric cohesion protector SGO1 were significantly downregulated in scc3 (Figure S11A). In contrast, the cohesin subunits SMC1 and SMC3, as well as the lateral/central elements PAIR3 and ZEP1, showed no differences in gene expression. The only upregulated axial element in scc3 spikelets was PAIR2.

We then investigated the expression levels of cohesin genes in axial/lateral element mutants (pair2 and pair3) (Figure S11B), and found that SMC3, SCC3 and SCC1 were significantly downregulated compared to the wild type. These findings suggest that SCC3 may thus be involved in the transcriptional regulation of meiosis genes, while axial elements affect cohesin proteins by regulating the expression levels of associated genes.

Discussion

Chromosome segregation during mitosis and meiosis encompasses a series of meticulously orchestrated events, including the formation of the cohesin complex, which serves diverse functions throughout the cell cycle (Moronta-Gines et al., 2019; Nasmyth and Haering, 2009). Here, we show SCC3 is a conserved cohesin subunit affecting sister chromatid assembly in both mitosis and meiosis. Crucially, our results also indicate SCC3 is a component of LE/AE that is required for homologous pairing and synapsis during meiosis, which is essential for the recombination process and CO formation.

Most research on the cohesin complex predominantly emphasizes its role during mitosis, attributed to the challenges associated with observing the morphology of sister chromatids throughout meiosis. In human cells, the separation of sister chromatids during mitosis occurs in the absence of the SCC3 homologs, SA1 and SA2. SA1 is indispensable for the cohesion of telomeres and chromosomal arms, while SA2 is crucial for centromeric cohesion (Canudas and Smith, 2009). In rice, SCC3 has only a single copy, therefore we observed all kinds of multifaceted defects in scc3. SCC3 is paramount not only for the stability of sister chromatid’s structure but also for facilitating the loading of cohesin. In yeast, the cohesin complex comprising SCC1, SMC1 and SMC3 remains intact in scc3. However, its chromosomal localization is impeded, most likely due to SCC3’s role in maintaining the stability of cohesion (Orgil et al., 2015; Pathania et al., 2021). Furthermore, SCC3 harbors a conserved DNA binding domain in the SCC3-SCC1 subcomplex, essential for the chromatin association of cohesins (Higashi et al., 2020). In yeast, the DNA-binding domain of the SCC3-SCC1 facilitates the recruitment of cohesin complexes to chromosomes, thus enhancing cohesin’s functionality during mitotic cell division (Kulemzina et al., 2012; Li et al., 2018a). These findings suggest that the interaction between SCC3 and SCC1 is vital for the stability of the cohesin ring and its continued association with DNA. In Arabidopsis, SCC3 is also involved in vegetative growth of plants, but its role in mitosis remains unclear (Chelysheva et al., 2005). Leveraging these antecedent insights, our investigation corroborates that SCC3 is integral to the initial formation of cohesion during mitosis. In the rice scc3 mutant, vegetative growth was severely inhibited. Our analyses indicate an aberration in the chromosomal structure during interphase in scc3 mutants, possibly due to SCC3’s necessity for stable chromosomal cohesin binding. Although an augmentation in the distance between sister chromatids was observed during prometaphase, the sister chromatids remained partially dissociated, suggesting the residual functionality of other cohesin subunits. Collectively, our findings underscore SCC3 as a conserved and pivotal cohesin protein within the mitotic process.

SCC3 not only functions in mitosis but also affects meiotic chromosome structure and function. In vertebrates, SCC3 is encoded by two paralogous genes, including the meiosis-specific subunit STAG3. Investigations utilizing SIM analysis on human stag3 have elucidated that chromosomal axes are regionally dissociated, and there is a substantial reduction in the synaptonemal complex (Biswas et al., 2016; Fukuda et al., 2014; Hopkins et al., 2014; Llano et al., 2012). These observations suggest STAG3 is required for synapsis between homologous chromosomes. In comparison, the morphological characteristics of meiocytes during meiosis are more discernible in plant cells. Specifically, in Arabidopsis scc3, a complete absence of fertility has been observed, underscoring the pivotal role of SCC3 in meiotic process, including unstable homologous pairing at the pachytene (Chelysheva et al., 2005). In analogous studies with other cohesin mutants, Atscc2 has been identified as essential for the recruitment of the meiosis-specific cohesin subunit REC8 (Wang et al., 2020). This suggests that mutations in cohesin-associated genes could potentially compromise the cohesin complex’s chromosome-binding capacity. Accordingly, in our investigation on meiosis, it has been determined that rice SCC3 facilitates the maintenance of the structure of sister chromatids during the preleptotene. More importantly, SCC3 is localized between homologous chromosomes as a critical AE, ensuring accurate homologous pairing and synapsis. SCC3’s presence is crucial for the normal localization of DMC1, RAD51, ZIP4 and HEI10, elements vital for progression of recombination and the formation of CO. Our proposed model suggests that SCC3 is a part of the cohesin ring, and functions in specific association with SCC1 (Figure S12A). Given SCC3’s multifaceted roles in cohesion dynamics and meiotic processes, we hypothesize that SCC3 may function as an axial element to stabilize loop boundaries in meiosis, thereby promoting the homologous pairing and synapsis (Figure S12B). Notably, it has been proposed that the facilitation of meiotic inter-homolog recombination is attributed to an inhibition of inter-sister-chromatid recombination, which is enforced by meiosis-specific constituents of the chromosome axis (Goldfarb and Lichten, 2010). In scc3 mutants, meiotic DSBs seemed to utilize sister chromatids as repair templates exclusively. Although essential meiotic proteins such as COM1, DMC1 and RAD51 were decreased in scc3, these proteins may still function in facilitating the inter-sister chromatid repair pathway, as DMC1 and RAD51 also involved in the repair process of inter-sister chromatids. We postulate that the presence of cohesin predisposes meiotic recombination towards an inter-homolog bias for repair. Conversely, in the absence of cohesin and when homologous recombination is compromised, recombination is biased towards inter-sister repair.

Cohesin is a multi-protein complex that plays an indispensable role in establishing sister chromatid cohesion and the high-order chromosome architecture in both somatic and germ cells. Distinct meiosis-specific subunits likely execute divergent functions (Cuadrado et al., 2015; Ishiguro, 2019; Schvarzstein et al., 2010). Our findings suggest that SCC3 and REC8 assume differential roles during mitosis and meiosis, respectively. REC8 was identified as a meiosis-specific protein, assuming the role of SCC1 in meiosis to constitute the cohesin complex. Our investigation reveals that although REC8 and SCC3 exhibit similar localization patterns, but they show many differences. Significantly, REC8 is completely absent in mitotic cells. Conversely, our data indicate that SCC3 is operative in both mitosis and meiosis, potentially playing a role in the loading of cohesins onto sister chromatids during interphase. It is noteworthy that the cohesin loading mechanism onto chromosomes during mitosis and meiosis remains consistent. Meiotic cohesin is loaded onto chromosomes during interphase, likely coinciding with the replication and meiotic fate acquisition of sporogenous cells. This indicates that REC8 does not participate in cohesion establishment at this juncture. Furthermore, we observed SCC3 signals at centromeres from diakinesis to metaphase I, whereas REC8 was undetectable at metaphase I. These distinct localization patterns further imply they may have divergent functional roles in meiosis. Previous research on Arabidopsis rec8 mutants highlighted chromosomes exhibiting a sticky phenotype that remained challenging to elucidate (Cai et al., 2003; Lambing et al., 2020). Similarly, rice rec8 meiocytes exhibited defective chromosomes with a pronounced sticky phenotype from diakinesis to metaphase I (Shao et al., 2011). In our study, chromosomes in scc3 did not display the stickiness observed in rec8 (Figure 6C). Moreover, the majority of meiotic proteins (including SCC3) failed to localize in rec8. This indicates the normal localization of SCC3 and other meiotic proteins depends on REC8, highlighting REC8 as a pivotal meiotic protein essential for the loading of other meiotic proteins. Additionally, we observed that the end processing of DSBs and the inter-sister/homolog strand invasion processes were profoundly impaired in rec8, leading us to speculate that the sticky chromosome phenotype in rec8 could be attributed to an unknown repair pathway. Our results underscore that SCC3 serves as a cohesin component and plays a significant role in cohesion establishment during both mitosis and meiosis. In contrast, the function of REC8 is only restricted to meiosis. While both proteins operate independently in mitosis, they are interrelated in meiosis, with REC8 emerging as a more critical factor in the meiotic process.

Material and methods

Meiotic chromosome preparation

Young rice panicles were fixed in Carnoy’s solution (ethanol: acetic acid, 3: 1, v/v). Microsporocytes at different meiotic stages were squashed in an acetocarmine solution on glass slides. Moreover, the slides were covered with a coverslip and washed with 45% acetic acid from each side. Subsequently, the slides were frozen in liquid nitrogen for 10 seconds, and then the coverslip was removed as soon as possible ensuring sufficient cells were fixed on the slide. After ethanol gradient dehydration (70%, 90% and 100%), the slides with chromosome spreads were counterstained with 4’,6-diamidino-2-phenylindole (DAPI, Vector Laboratories, Burlingame, CA, USA) solution. Chromosome behaviors were observed and captured using a ZEISS A2 microscope imaging system with a micro-charge-coupled (micro CCD) device camera.

Antibody production

To generate the antibody against SCC3, a 648 bp C-terminal fragment of SCC3 complementary DNA (amino acids 910-1116) was amplified using primers SCC3-Ab-F and SCC3-Ab-R (Table S1). The PCR product was cloned into the expression vector pET-30a (Novagen, Madison, WI, USA). The fusion peptide expression and purification were carried out as described previously (Miao et al., 2021). The polyclonal antibody was raised from mouse and rabbit. Other antibodies were generated in our laboratory previously.

Immunofluorescence assay

Fresh young panicles were fixed in 4% (w/v) paraformaldehyde for 30 min at room temperature. Different meiotic stages of anthers were squashed in a drop of 1×PBS solution on glass slides. After freezing in liquid nitrogen and ethanol dehydration, slides were incubated with different combinations of diluted antibodies (1:100) in a humid chamber at 37 °C for 2 h. Thereafter, the slides were washed with 1×PBS solution three times and were further incubated for 1 h with the appropriate fluorochrome-coupled secondary antibody, including fluorescein-isothiocyanate-conjugated goat anti-mouse antibody (Southern Biotech, Birmingham, AL, USA), rhodamine-conjugated goat anti-rabbit antibody (Southern Biotech), and AMCA-conjugated goat anti-guinea pig antibody (Jackson ImmunoResearch, West Grove, PA, USA). Slides were washed with 1×PBS solution three times and eventually stained with DAPI. The immunofluorescence signals were analyzed and captured using a ZEISS A2 microscope imaging system.

Fluorescence in situ hybridization

Young panicles of both wild-type and mutants were fixated in Carnoy’s solution (ethanol: acetic acid, 3: 1, v/v). The fluorescence in situ hybridization (FISH) analysis was performed as a detailed protocol previously (Cheng, 2013). The pAtT4 clone containing telomere repeats, the pTa794 clone containing 5S ribosomal RNA genes, and the bulked oligonucleotide probes specific to chromosome 12 were used as probes in FISH analysis. Rhodamine anti-digoxigenin was used for digoxigenin-labeled probes. Chromosome images were captured using an Olympus (Shinjuku-ku, Tokyo, Japan) BX51 fluorescence microscope with a micro CCD camera using software IPLAB4.0 (BD Biosciences, San Jose, CA, USA).

Yeast two-hybrid assay

The yeast two-hybrid (Y2H) assays were conducted by the full-length CDSs of RAD21.1, SCC3, REC8, SMC1 and SMC3 independently cloned into the pGBKT7 or pGADT7 vector. Co-transformants were selected on SD/DDO (SD-Leu-Trp) medium at 30 °C for 3-4 days. Positive interactions of transformants were selected on SD/QDO (SD-Leu-Trp-His-Ade) medium containing aureobasidin A. All primers used to construct plasmids are listed in Table S1. The detailed protocol was described in the manufacturer’s handbook (Yeast protocols handbook; PT3024-1; Clontech).

BiFC assay

To conduct BiFC assays, SCC3 and RAD21.1 were amplified by KOD-plus polymerase and ligated into BiFC vectors, including pSCYNE (SCN) and pSCYCE (SCC). The constructed plasmids were transformed into protoplasts extracted from the young stem of rice etiolated seedlings. After incubation in the dark for 18 h at 28°C, the CFP signals were captured under a confocal laser scanning microscope at an excitation wavelength 405 nm (Leica TCS SP5, Wetzlar, Germany).

Super-resolution structured illumination microscopy and image analysis

Immunofluorescence assays were conducted by the procedures described above. Meiotic cells were scanned under the DeltaVision microscope (OMXTM V4; GEhealthcare, Chicago, IL, USA) and the super-resolution images were processed with SoftWoRx (Applied Precision, Issaquah, WA, USA).

RNA extraction and real-time PCR

Real-time PCR assays were used to detect the expression patterns of SCC3. Total RNA was extracted from different tissues in wild-type including root, stem, leaf, sheath, spikelet, 2 cm long panicle, 2-3 cm long panicle, 3-5 cm long panicle, 5-7 cm long panicle and 7 cm long panicle using TRIzol reagent. After digestion with RNase-free DNaseI to remove genomic DNA, reverse transcription PCR was performed to synthesize cDNA using SuperScriptTM III Reverse Transcriptase by manufacturer’s protocol (Invitrogen). Quantitative real-time PCR assays were performed on the Bio-Rad CFX96 instrument using Evagreen (Biotium, Freemont, CA, USA) with a procedure of 98 °C 3min, 40 cycles of 98 °C 15s and 60 °C 20s. The UBIQUITIN gene was used as an internal reference. Three biological repeats were performed for each sample. Primers used for real-time PCR assays were listed in Table S1.

Phylogenetic analysis

The full-length amino acid sequence of SCC3 was selected as a template and searched by the NCBI blastp tool to identify similar proteins among species. A detailed protocol for further analysis has been reported previously (Li et al., 2018b; Ren et al., 2018; Ren et al., 2020). Filtered sequences were downloaded and used for constructing neighbor-joining trees by MEGA5 software and processed with EvolvieW (https://evolgenius.info/evolview/).

Multiple sequence alignments

Multiple alignments were performed using the online toolkit MAFFT (https://toolkit.tuebingen.mpg.de/mafft) and processed with ESPRIPT3 (http://espript.ibcp.fr/ESPript/ESPript/) by previously detailed reported (Zhao et al., 2021).

Experimental model and subject details

The rice scc3 mutant was produced by CRISPR-Cas9 toolkit in japonica rice variety Yandao 8. Primers used for CRISPR-Cas9 are listed in Table S1. Other meiosis mutants pair3 (Wuxiangjing 9), pair2 (Nipponbare), zep1 (Nipponbare), rec8 (Yandao 8), sgo1 (Yandao 8), spo11-1 (Guangluai 4), dmc1 (Nipponbare) have been previously described (Wang et al., 2016). Yandao 8 was used as the wild-type. All the rice plants were grown in a paddy field under natural rice-growing conditions.

Quantification and statistical analysis

At least three independent biological repeats were performed for each experiment. Values in Figures are means ± SD. The significant difference between the wild type and corresponding mutants was evaluated by two-tailed Student’s t-test and p < 0.05 was considered as a criterion for judging statistically significant difference. The significance labeled in the Figures represents the following meanings: *, p < 0.05; **, p < 0.01; ***, p < 0.001; ****, p < 0.0001.

Supporting information